---
title: "Meta-analysis with `glmmTMB` package `equalto` covariance structure"
author: "Coralie Williams"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    page-layout: full
    toc: true
    toc-location: left
    toc-depth: 2
    theme: default
    embed-resources: true
    code-fold: false
    code-tools: true
    number-sections: true
editor_options: 
  chunk_output_type: console
---

# Overview

This webpage presents the `glmmTMB` package `equalto` covariance structure to fit a wide range of meta-analytical models.

Let's load the packages needed. 

```{r}
# load packages
library(MASS)
library(metafor)
library(nlme)
library(kableExtra)
library(insight) #to get degrees of freedom
library(microbenchmark) #for evaluating speed of functions
```

Install the branch of `glmmTMB` with `equalto`. If you have `glmmTMB` already installed, remove this version and reload R afterwards (Ctrl+Shift+F10). Then we will build `glmmTMB` from the following branch:  [https://github.com/coraliewilliams/glmmTMB/tree/equalto_covstruc](https://github.com/coraliewilliams/glmmTMB/tree/equalto_covstruc)

```{r}
#| eval: FALSE
#| output: false
#| warnings: false

# remove glmmTMB (if installed)
remove.packages("glmmTMB") 

# build package from branch: https://github.com/coraliewilliams/glmmTMB/tree/equalto_covstruc
remotes::install_github("coraliewilliams/glmmTMB", ref="equalto_covstruc", subdir="glmmTMB")
```

Now load `glmmTMB` and check if `equalto` is there:

```{r}
# load it
library(glmmTMB)

# check the available covstruc (you should see equalto in number 13)
(glmmTMB:::.valid_covstruct)

```



# Multilevel meta-analysis

We want to fit a multilevel meta-analysis model that synthesises effect sizes across studies by accounting for variability at the study level, at the observation (effect size) level, and by modelling correlated residual errors. 

We specify the model as 
$$
y_{ij} = b_0 + u_i + m_{ij} + e_{ij},
$$

where $i$ indexes studies and $j$ indexes effect sizes within study $i$. Here, $b_0$ is the overall mean effect, $u_i \sim N(0, s2\_u.I)$ represents the study-level random effect, $m_{ij} \sim N(0, s2\_m.I)$ captures the observation-level variability, and $e_{ij} \sim N(0, \text{VCV})$ are the residual errors, with $\text{VCV}$ representing a block-diagonal variance-covariance matrix that accounts for correlations among effect sizes from the same study.



### Simulate data

Now let's simulation some data following the structure of this model. 

```{r}
# set up indices
set.seed(123)
k.studies <- 50 # number of studies
k.per.study <- 10 # number of effect sizes per study
study <- rep(seq_len(k.studies), times=k.per.study) # study id 
k <- length(study) # total number of effect sizes
id <- seq_len(k) # id of between study effect sizes 
es.id <- unlist(lapply(k.per.study, seq_len)) # id of within study effect sizes


# fixed effect coeff true value (i.e. the overall mean)
b0 <- 0

# simulate sampling errors variances
vi <- rbeta(k, 2, 20)

# get VCV matrix assuming within-study correlation of rho=0.5 
VCV <- matrix(0, nrow = k, ncol = k) 
for (i in 2:k) {
  for (j in 1:i) {
    if (study[i] == study[j]) {
      VCV[i,j] <- 0.5 * sqrt(vi[i] * vi[j]) 
    }
  }
}
VCV[upper.tri(VCV)] <- t(VCV)[upper.tri(VCV)] # fill in upper diagonal
diag(VCV) <- vi # fill in diagonal


# random effects
sigma2.u <- 0.2
sigma2.m <- 0.3
u <- rnorm(k.studies, 0, sqrt(sigma2.u))[study]
m <- rnorm(k, 0, sqrt(sigma2.m))
#e <- rnorm(k, 0, sqrt(vi))[study] #without within-study correlation (diag VCV)
e <- mvrnorm(n = 1, mu = rep(0, length(vi)), Sigma = VCV) #excluding within-study correlation


# compute y
y <- b0 + u + m + e


# combine into dataframe
dat <- data.frame(y = y,
                  vi = vi,
                  study = study,
                  id = id,
                  es.id = es.id,
                  obs = as.factor(id), # unique ID for propto and equalto (really important it is a factor variable)
                  g = 1 # group ID for propto and equalto
)
```


### Run model

Before we fit the glmmTMB model we need to obtain the variance covariance (VCV) matrix of the sampling errors. 

```{r}
# add code using vcalc just for demonstration
```

Now fit the glmmTMB model using equato
*** add more info about what each part is


```{r}
#Fit glmmTMB model with equalto --------------------------------------------
fit.equalto <- glmmTMB(y ~ 1 + (1|study) + equalto(0 + obs|g,VCV),
                       data=dat,
                       REML=T)
summary(fit.equalto)
```


Let's compare with metafor
*** add more info about what each part is in comparison to glmmTMB


```{r}
# Fit metafor model
fit.metafor <- rma.mv(y, VCV,
                      random = list(~1 | study, ~1 | id),
                      test = "t",
                      dfs = "contain",
                      data=dat) #used REML by default

##### Compare output 
metafor <- data.frame(model = "metafor", 
                      est = fit.metafor$b[[1]], 
                      se = fit.metafor$se[[1]], 
                      zval = fit.metafor$zval,
                      pval = fit.metafor$pval,
                      sigma.u = sqrt(fit.metafor$sigma2[1]),
                      sigma.s = sqrt(fit.metafor$sigma2[2]),
                      sigma.total = sum(sqrt((fit.metafor$sigma2))),
                      logLik = logLik(fit.metafor)[1],
                      df = attr(logLik(fit.metafor), "df"), ## model-based 
                      df.res = df.residual(fit.metafor)) ## residual df

equalto <- data.frame(model = "equalto",
                      est = unlist(fixef(fit.equalto))[[1]],
                      se = as.numeric(sqrt(vcov(fit.equalto)[[1]])),
                      zval = summary(fit.equalto)$coefficients$cond[3],
                      pval = summary(fit.equalto)$coefficients$cond[3],
                      sigma.u = exp(fit.equalto$fit$par[[2]]),
                      sigma.s = sigma(fit.equalto),
                      sigma.total = sum(exp(fit.equalto$fit$par)),
                      logLik = logLik(fit.equalto),
                      df = attr(logLik(fit.equalto), "df"), ## model-based df
                      df.res = df.residual(fit.equalto)) ## residual df


output <- rbind(metafor, equalto)
kable(output)

```



# Comparison of `glmmTMB` with `metafor` and `nlme` R packages (using [this blog post](https://jepusto.com/posts/Bug-in-nlme-with-fixed-sigma/))

Let's compare the model estimates across two other packages. The following code is inspired from the post by J.E.Pustejovsky: [https://jepusto.com/posts/Bug-in-nlme-with-fixed-sigma/](https://jepusto.com/posts/Bug-in-nlme-with-fixed-sigma/)


## Basic random effect

```{r}
bcg_example <- function(method = "REML", constant_var = FALSE) {
  
  data(dat.bcg)
  dat <- escalc(measure="OR", ai=tpos, bi=tneg, ci=cpos, di=cneg, data=dat.bcg)
  dat$g <- 1
  dat$trial <- as.factor(dat$trial)
  
  v_bar <- mean(dat$vi)
  if (constant_var) dat$vi <- v_bar
  
  # random-effects model using rma.uni()
  LOR_uni_fit <- rma(yi, vi, data=dat, method = method)
  LOR_uni <- with(LOR_uni_fit, 
                  data.frame(f = "rma.uni", 
                             logLik = logLik(LOR_uni_fit),
                             df = attr(logLik(LOR_uni_fit), "df"),
                             est = as.numeric(b), 
                             se = se, 
                             tau = sqrt(tau2)))
  
  # random-effects model using rma.mv()
  LOR_mv_fit <- rma.mv(yi, vi, random = ~ 1 | trial, data=dat, method = method)
  LOR_mv <- with(LOR_mv_fit, 
                 data.frame(f = "rma.mv", 
                            logLik = logLik(LOR_mv_fit),
                            df = attr(logLik(LOR_mv_fit), "df"),#model-based degrees of freedom, i.e. the number of (estimated) parameters
                            est = as.numeric(b), 
                            se = se, 
                            tau = sqrt(sigma2)))
  
  # random-effects model using lme()
  if (constant_var) {
    LOR_lme_fit <- lme(yi ~ 1, data = dat, method = method, 
                       random = ~ 1 | trial,
                       control = lmeControl(sigma = sqrt(v_bar)))
    tau <- sqrt(as.numeric(coef(LOR_lme_fit$modelStruct$reStruct, unconstrained = FALSE)) * v_bar) 
  } else {
    LOR_lme_fit <- lme(yi ~ 1, data = dat, method = method, 
                       random = ~ 1 | trial,
                       weights = varFixed(~ vi),
                       control = lmeControl(sigma = 1))
    tau <- sqrt(as.numeric(coef(LOR_lme_fit$modelStruct$reStruct, unconstrained = FALSE)))
  }
  LOR_lme <- data.frame(f = "lme", 
                        logLik = logLik(LOR_lme_fit),
                        df = attr(logLik(LOR_lme_fit), "df"),  
                        est = as.numeric(fixef(LOR_lme_fit)), 
                        se = as.numeric(sqrt(vcov(LOR_lme_fit))), 
                        tau = tau)
  
  
  # random-effects model using glmmTMB() and equalto
  if (method=="REML") REML <- TRUE else REML <- FALSE
  
  if (constant_var) {
    VCV <- diag(v_bar, nrow = nrow(dat))
    LOR_glmm_fit <- glmmTMB(yi ~ 1 + equalto(0 + trial|g,VCV),
                            REML = REML,
                            data=dat)
    tau <- sigma(LOR_glmm_fit) 
  } else {
    VCV <- diag(dat$vi)
    LOR_glmm_fit <- glmmTMB(yi ~ 1 + equalto(0 + trial|g,VCV),
                            REML = REML,
                            data=dat)
    tau <- sigma(LOR_glmm_fit) 
  }
  LOR_glmm <- data.frame(f = "glmmTMB", 
                        logLik = logLik(LOR_glmm_fit)[1],
                        df = attr(logLik(LOR_glmm_fit), "df"),                         
                        est = as.numeric(unlist(fixef(LOR_glmm_fit))[[1]]), 
                        se = as.numeric(sqrt(vcov(LOR_glmm_fit)[[1]])), 
                        tau = tau)
  
  
  kable(rbind(LOR_uni, LOR_mv, LOR_glmm, LOR_lme))
  
}

bcg_example("REML", constant_var = FALSE)
bcg_example("REML", constant_var = TRUE)

bcg_example("ML", constant_var = FALSE)
bcg_example("ML", constant_var = TRUE)

```


## Bi-variate random effects model


```{r}

bcg_bivariate <- function(method = "REML", constant_var = FALSE) {
  
  data(dat.bcg)
  dat_long <- to.long(measure="OR", ai=tpos, bi=tneg, ci=cpos, di=cneg, data=dat.bcg)
  levels(dat_long$group) <- c("exp", "con")
  dat_long$group <- relevel(dat_long$group, ref="con")
  dat_long$g <- 1
  dat_long$ID <- as.factor(1:nrow(dat_long))
  dat_long <- escalc(measure="PLO", xi=out1, mi=out2, data=dat_long)
  
  v_bar <- mean(dat_long$vi)
  if (constant_var) dat_long$vi <- v_bar
  
  
  # bivariate random-effects model using rma.mv()
  bv_rma_fit <- rma.mv(yi, vi, mods = ~ group, 
                       random = ~ group | study, 
                       struct = "UN", method = method,
                       data=dat_long)
  bv_rma <- with(bv_rma_fit, data.frame(f = "rma.mv",
                                        logLik = logLik(bv_rma_fit),
                                        df = attr(logLik(bv_rma_fit), "df"), 
                                        est1 = as.numeric(b)[1], 
                                        se1 = se[1], 
                                        est2 = as.numeric(b)[2],
                                        se2 = se[2],
                                        tau1 = sqrt(tau2[1]),
                                        tau2 = sqrt(tau2[2])))
  
  # bivariate random-effects model using lme()
  if (constant_var) {
    bv_lme_fit <- lme(yi ~ group, data = dat_long, method = method, 
                      random = ~ group | study,
                      control = lmeControl(sigma = sqrt(v_bar)))
    tau_sq <- colSums(coef(bv_lme_fit$modelStruct$reStruct, unconstrained = FALSE) * matrix(c(1,0,0, 1,2,1), 3, 2)) * v_bar
    
  } else {
    bv_lme_fit <- lme(yi ~ group, data = dat_long, method = method, 
                      random = ~ group | study,
                      weights = varFixed(~ vi),
                      control = lmeControl(sigma = 1))
    tau_sq <- colSums(coef(bv_lme_fit$modelStruct$reStruct, unconstrained = FALSE) * matrix(c(1,0,0, 1,2,1), 3, 2))
  }
  bv_lme <- data.frame(f = "lme",
                       logLik = logLik(bv_lme_fit),
                       df = attr(logLik(bv_lme_fit), "df"), 
                       est1 = as.numeric(fixef(bv_lme_fit))[1],
                       se1 = as.numeric(sqrt(diag(vcov(bv_lme_fit))))[1],
                       est2 = as.numeric(fixef(bv_lme_fit))[2],
                       se2 = as.numeric(sqrt(diag(vcov(bv_lme_fit))))[2],
                       tau1 = sqrt(tau_sq[1]),
                       tau2 = sqrt(tau_sq[2]))
  
  
  
  # bivariate random-effects model using glmmTMB() and equalto
  if (method=="REML") REML <- TRUE else REML <- FALSE
  
  if (constant_var) {
    VCV <- diag(v_bar, nrow = nrow(dat_long))
    bv_glmm_fit <- glmmTMB(yi ~ group + (group|study) + equalto(0 + ID|g,VCV), 
                            REML = REML,
                            data=dat_long)
    tau <- c(exp(bv_glmm_fit$fit$par)[2:3]) 
    ## 1st theta is random intercept study(intercept), 2nd is random slope study(groupexp), and 3rd is the covariance term 

  } else {
    VCV <- diag(dat_long$vi)
    bv_glmm_fit <- glmmTMB(yi ~ group + (group|study) + equalto(0 + ID|g,VCV),
                            REML = REML,
                            data=dat_long)
    tau <- c(exp(bv_glmm_fit$fit$par)[2:3]) 
  }
  bv_glmm <- data.frame(f = "glmmTMB", 
                        logLik = logLik(bv_glmm_fit)[1],
                        df = attr(logLik(bv_glmm_fit), "df"), 
                        est1 = as.numeric(unlist(fixef(bv_glmm_fit))[[1]]), 
                        se1 = as.numeric(sqrt(diag(vcov(bv_glmm_fit)[[1]])))[1], 
                        est2 = as.numeric(unlist(fixef(bv_glmm_fit))[[2]]), 
                        se2 = as.numeric(sqrt(diag(vcov(bv_glmm_fit)[[1]])))[2], 
                        tau1 = tau[1],
                        tau2 = tau[2])
  
  
  
  kable(rbind(bv_rma, bv_glmm, bv_lme))
  
}

bcg_bivariate("REML", constant_var = FALSE)
bcg_bivariate("REML", constant_var = TRUE)

bcg_bivariate("ML", constant_var = FALSE) ##non-convergence for glmmTMB
bcg_bivariate("ML", constant_var = TRUE) ##very different tau1 and ta2 for glmmTMB

```





# Speed comparison

```{r}
# for propto
colnames(VCV) <- 1:nrow(dat)
row.names(VCV) <- 1:nrow(dat)

# using the data from the second simulation - this runs 100 times by default
bench_fit <- microbenchmark(
  metafor = rma.mv(y, VCV, random = list(~1 | study, ~1 | id),
                   test = "t",
                   dfs = "contain",
                   data=dat),
  equalto = glmmTMB(y ~ 1 + (1|study) + equalto(0 + obs|g,VCV),
                    data=dat,
                    REML=T),
  propto = glmmTMB(y ~ 1 + (1|study) + propto(0 + obs|g,VCV), # expect it to be the same
                    data=dat,
                    REML=T)
)

bench_fit
plot(bench_fit)
```

Note: BLAS is installed on my session and seems to improve the performance of metafor a lot more. 



# Next steps....

Alternative strategy is to use Gaussian Markov Random Field in TMB cpp that uses the inverse of the covariance matrix.

`density::GMRF_t`
https://kaskr.github.io/adcomp/classdensity_1_1GMRF__t.html


# Session info 

```{r}
#| label: Reproducibility-SessionInfo-R-environment
#| fig-align: "center"
#| out-width: '100%'
#| results: asis
#| message: false
#| warnings: false

#library(sessioninfo)
#library(details)

#si <- session_info()
#si$packages <- si$packages %>% 
#  filter(package %in% c("metafor", "glmmTMB", "TMB", "Matrix", "corpcor", "dplyr", "kableExtra", "lattice"))

#details(si, summary = 'Current session info', open = FALSE)

```