---
title: "Meta-analysis with `glmmTMB` package `equalto` covariance structure"
author: "Coralie Williams"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    page-layout: full
    toc: true
    toc-location: left
    toc-depth: 2
    theme: default
    embed-resources: true
    code-fold: false
    code-tools: true
    number-sections: true
editor_options: 
  chunk_output_type: console
---

# Overview

This webpage describes the `equalto` covariance structure within the `glmmTMB` package to fit a wide range of meta-analytical models.

First, let's load the packages needed.

```{r}
#| message: false
#| warnings: false

# load packages
library(MASS)
library(metafor)
library(nlme)
library(kableExtra, verbose=F)
library(insight) #to get degrees of freedom
library(microbenchmark) #for evaluating speed of functions
```

We will install `glmmTMB` using the following branch branch: <https://github.com/coraliewilliams/glmmTMB/tree/equalto_covstruc>. 
<!-- (If you have `glmmTMB` already installed, remove this version and reload R afterwards with this shortcut: Ctrl+Shift+F10) -->

```{r}
#| eval: FALSE
#| output: false
#| warnings: false

# remove glmmTMB (if installed) and reload R
remove.packages("glmmTMB") 

# build package from branch: https://github.com/coraliewilliams/glmmTMB/tree/equalto_covstruc
remotes::install_github("coraliewilliams/glmmTMB", ref="equalto_covstruc", subdir="glmmTMB")
```

Check if `equalto` is there:

```{r}
# load it
library(glmmTMB)

# check the available covstruc (you should see equalto in number 13)
(glmmTMB:::.valid_covstruct)

```

<!-- Yes? Great - let's move on to fitting some meta-analytical models.  -->

# Multilevel meta-analysis

We are interested in fitting a multilevel meta-analysis model that synthesises effect sizes across studies by accounting for variability at the study level, at the observation (effect size) level, and by modelling correlated within-study residual errors.

We specify the model as 
$$
y_{ij} = b_0 + u_i + m_{ij} + e_{ij},
$$

where $i$ indexes studies and $j$ indexes effect sizes within study $i$. Here, $b_0$ is the overall mean effect, $u_i \sim N(0, \sigma^2_{u}I)$ represents the among-study random effect, $m_{ij} \sim N(0, \sigma^2_{m}I)$ captures the within-study random effect (at the observation-level), and $e_{ij} \sim N(0, \text{VCV})$ are the sampling errors at the observation-level. $\text{VCV}$ represents a block-diagonal variance-covariance matrix including known (or calculated) sampling variances $\sigma^2_{e}=v_i$ on the diagonal and accounts for correlations among sampling errors from the same study.


### Simulate some data

Now let's simulation data following the structure of the above multilevel meta-analysis.

```{r}
# set up indices
set.seed(123)
k.studies <- 50 # number of studies
k.per.study <- 10 # number of effect sizes per study
study <- rep(seq_len(k.studies), times=k.per.study) # study id 
k <- length(study) # total number of effect sizes
id <- seq_len(k) # id of between study effect sizes 
es.id <- unlist(lapply(k.per.study, seq_len)) # id of within study effect sizes


# fixed effect coeff true value (i.e. the overall mean)
b0 <- 0

# simulate sampling errors variances
vi <- rbeta(k, 2, 20)

# get VCV matrix assuming within-study correlation of rho=0.5 
V <- matrix(0, nrow = k, ncol = k) 
for (i in 2:k) {
  for (j in 1:i) {
    if (study[i] == study[j]) {
      V[i,j] <- 0.5 * sqrt(vi[i] * vi[j]) 
    }
  }
}
V[upper.tri(V)] <- t(V)[upper.tri(V)] # fill in upper diagonal
diag(V) <- vi # fill in diagonal


# random effects
sigma2.u <- 0.2
sigma2.m <- 0.3
u <- rnorm(k.studies, 0, sqrt(sigma2.u))[study]
m <- rnorm(k, 0, sqrt(sigma2.m))
#e <- rnorm(k, 0, sqrt(vi))[study] #without within-study correlation (diag VCV)
e <- mvrnorm(n = 1, mu = rep(0, length(vi)), Sigma = V) 


# compute y
y <- b0 + u + m + e


# combine into dataframe
dat <- data.frame(y = y,
                  vi = vi,
                  study = study,
                  id = id,
                  es.id = es.id,
                  obs = as.factor(id), #unique ID for propto and equalto (really important it is a factor variable)
                  g = 1 #group ID for propto and equalto (a constant here)
)
```

### Run model

Before we fit the glmmTMB model we need to obtain the within-study error variance-covariance matrix (which accounts for within-study correlation) assuming a constant correlation of $\rho=0.5$. The correlation between effect sizes from different studies is assumed to be zero.

```{r}
VCV <- vcalc(vi, cluster=study, obs=obs, data=dat, rho=0.5)
```

Now we fit the glmmTMB model. We specify the study-level random effect $u_i$ as `(1|study)`, the sampling errors $e_{ij}$ are specified using the covariance structure `equalto(0 + obs|g,VCV)` allowing to specify the known/calculated VCV matrix. The $m_{ij}$ within-study random effect that captures the variability at the observation level is modelled as the residual error in this model. 

```{r}
#Fit glmmTMB model with equalto --------------------------------------------
fit.equalto <- glmmTMB(y ~ 1 + (1|study) + equalto(0 + obs|g,VCV),
                       data=dat,
                       REML=T)
```

Let's compare with the `metafor` package `rma.mv` function that fits a multilevel model with random effects. Here the model is specified with two random effects: `(1|study)` and `(1|id)`, corresponding to the among-study ($u_i$) and within-study ($m_{ij}$) random effects, respectively. The sampling errors $e_{ij}$ VCV matrix is specified using the `V` argument.

```{r}
# Fit metafor model
fit.metafor <- rma.mv(y, VCV,
                      random = list(~1 | study, ~1 | id),
                      test = "t",
                      dfs = "contain",
                      data=dat) #used REML by default

##### Compare output 
metafor <- data.frame(model = "metafor", 
                      est = fit.metafor$b[[1]], 
                      se = fit.metafor$se[[1]], 
                      zval = fit.metafor$zval,
                      pval = fit.metafor$pval,
                      sigma.u = sqrt(fit.metafor$sigma2[1]), ## among study variance
                      sigma.m = sqrt(fit.metafor$sigma2[2]), ## withins study variance
                      sigma.total = sum(sqrt((fit.metafor$sigma2))),
                      logLik = logLik(fit.metafor)[1],
                      df = attr(logLik(fit.metafor), "df"), ## model-based 
                      df.res = df.residual(fit.metafor)) ## residual df

equalto <- data.frame(model = "equalto",
                      est = unlist(fixef(fit.equalto))[[1]],
                      se = as.numeric(sqrt(vcov(fit.equalto)[[1]])),
                      zval = summary(fit.equalto)$coefficients$cond[3],
                      pval = summary(fit.equalto)$coefficients$cond[4],
                      sigma.u = exp(fit.equalto$fit$par[[2]]), ##among study variance estimate
                      sigma.m = sigma(fit.equalto), ##within study variance estimate
                      sigma.total = sum(exp(fit.equalto$fit$par)),
                      logLik = logLik(fit.equalto),
                      df = attr(logLik(fit.equalto), "df"), ## model-based df
                      df.res = df.residual(fit.equalto)) ## residual df


output <- rbind(metafor, equalto)
kable(output)

```

The overall mean estimate and random effect variance estimates are identical across the two packages :-D To note though: the standard errors differ to the fourth decimal point (and hence also the z-value and p-value), as well as the log-likelihoods and residual degrees of freedom (which is expected?).



# Comparison of `glmmTMB` with `metafor` and `nlme` R packages (using code from [this blog post](https://jepusto.com/posts/Bug-in-nlme-with-fixed-sigma/))

Let's compare the model estimates across two other packages. The following code has been taken from the post by J.E.Pustejovsky.

Here we will fit two types of models and compare results across different R packages: a basic random effect model and a three-level random effect model using `metafor`, `nlme`, and `glmmTMB`. We will store the model estimates, standard errors, degrees of freedom and log-likelihoods in a data frame for comparison. We also store $\tau$ which is the estimated standard deviation of the underlying true effect sizes across studies (also called the between-study heterogeneity).

## Basic random effect

```{r}
bcg_example <- function(method = "REML", constant_var = FALSE) {
  
  data(dat.bcg)
  dat <- escalc(measure="OR", ai=tpos, bi=tneg, ci=cpos, di=cneg, data=dat.bcg)
  dat$g <- 1
  dat$trial <- as.factor(dat$trial)
  
  v_bar <- mean(dat$vi)
  if (constant_var) dat$vi <- v_bar
  
  # random-effects model using rma.uni()
  LOR_uni_fit <- rma(yi, vi, data=dat, method = method)
  LOR_uni <- with(LOR_uni_fit, 
                  data.frame(f = "rma.uni", 
                             logLik = logLik(LOR_uni_fit),
                             df = attr(logLik(LOR_uni_fit), "df"),
                             est = as.numeric(b), 
                             se = se, 
                             tau = sqrt(tau2)))
  
  # random-effects model using rma.mv()
  LOR_mv_fit <- rma.mv(yi, vi, random = ~ 1 | trial, data=dat, method = method)
  LOR_mv <- with(LOR_mv_fit, 
                 data.frame(f = "rma.mv", 
                            logLik = logLik(LOR_mv_fit),
                            df = attr(logLik(LOR_mv_fit), "df"), #model-based dfs, i.e. the number of (estimated) params
                            est = as.numeric(b), 
                            se = se, 
                            tau = sqrt(sigma2)))
  
  # random-effects model using lme()
  if (constant_var) {
    LOR_lme_fit <- lme(yi ~ 1, data = dat, method = method, 
                       random = ~ 1 | trial,
                       control = lmeControl(sigma = sqrt(v_bar)))
    tau <- sqrt(as.numeric(coef(LOR_lme_fit$modelStruct$reStruct, unconstrained = FALSE)) * v_bar) 
  } else {
    LOR_lme_fit <- lme(yi ~ 1, data = dat, method = method, 
                       random = ~ 1 | trial,
                       weights = varFixed(~ vi),
                       control = lmeControl(sigma = 1))
    tau <- sqrt(as.numeric(coef(LOR_lme_fit$modelStruct$reStruct, unconstrained = FALSE)))
  }
  LOR_lme <- data.frame(f = "lme", 
                        logLik = logLik(LOR_lme_fit),
                        df = attr(logLik(LOR_lme_fit), "df"),  
                        est = as.numeric(fixef(LOR_lme_fit)), 
                        se = as.numeric(sqrt(vcov(LOR_lme_fit))), 
                        tau = tau)
  
  
  # random-effects model using glmmTMB() and equalto
  if (method=="REML") REML <- TRUE else REML <- FALSE
  
  if (constant_var) {
    VCV <- diag(v_bar, nrow = nrow(dat))
    LOR_glmm_fit <- glmmTMB(yi ~ 1 + equalto(0 + trial|g,VCV),
                            REML = REML,
                            data=dat)
    tau <- sigma(LOR_glmm_fit) 
  } else {
    VCV <- diag(dat$vi)
    LOR_glmm_fit <- glmmTMB(yi ~ 1 + equalto(0 + trial|g,VCV),
                            REML = REML,
                            data=dat)
    tau <- sigma(LOR_glmm_fit) 
  }
  LOR_glmm <- data.frame(f = "glmmTMB", 
                        logLik = logLik(LOR_glmm_fit)[1],
                        df = attr(logLik(LOR_glmm_fit), "df"),                         
                        est = as.numeric(unlist(fixef(LOR_glmm_fit))[[1]]), 
                        se = as.numeric(sqrt(vcov(LOR_glmm_fit)[[1]])), 
                        tau = tau)
  
  
  kable(rbind(LOR_uni, LOR_mv, LOR_glmm, LOR_lme))
  
}

bcg_example("REML", constant_var = FALSE)
bcg_example("REML", constant_var = TRUE)

bcg_example("ML", constant_var = FALSE)
bcg_example("ML", constant_var = TRUE)

```

Here we see that glmmTMB is almost identical to `metafor` (assumed to be the standard results). A few things to note:

-   The log likelihood is different for glmmTMB when using `REML` (but it is the same as metafor when using`ML`).
-   The standard error is different for glmmTMB to the fourth decimal point when the variance is not constant (`constant_var=FALSE`), but exactly the same when the variance is constant (`constant_var=TRUE`). 



## Three-level random-effects model

```{r}
Konstantopoulos <- function(method = "REML", constant_var = FALSE) {
  
  dat <- get(data(dat.konstantopoulos2011))
  v_bar <- mean(dat$vi)
  dat$g <- 1
  dat$obs <- as.factor(1:nrow(dat))
  if (constant_var) dat$vi <- v_bar
  
  # multilevel random-effects model using rma.mv()
  ml_rma_fit <- rma.mv(yi, vi, random = ~ 1 | district/school, data=dat, method = method)
  
  ml_rma <- with(ml_rma_fit, 
                 data.frame(f = "rma.mv", 
                            logLik = logLik(ml_rma_fit),
                            df = attr(logLik(ml_rma_fit), "df"), 
                            est = as.numeric(b), 
                            se = se, 
                            tau1 = sqrt(sigma2[1]), 
                            tau2 = sqrt(sigma2[2])))
  
  # multilevel random-effects model using lme()
  if (constant_var) {
    ml_lme_fit <- lme(yi ~ 1, data = dat, method = method, 
                      random = ~ 1 | district / school,
                      control = lmeControl(sigma = sqrt(v_bar)))
    tau <- sqrt(as.numeric(coef(ml_lme_fit$modelStruct$reStruct, unconstrained = FALSE)) * v_bar)
    
  } else {
    ml_lme_fit <- lme(yi ~ 1, data = dat, method = method, 
                      random = ~ 1 | district / school,
                      weights = varFixed(~ vi),
                      control = lmeControl(sigma = 1))
    tau <- sqrt(as.numeric(coef(ml_lme_fit$modelStruct$reStruct, unconstrained = FALSE)))
    
  }  
  ml_lme <- data.frame(f = "lme",
                       logLik = logLik(ml_lme_fit),
                       df = attr(logLik(ml_lme_fit), "df"), 
                       est = as.numeric(fixef(ml_lme_fit)),
                       se = as.numeric(sqrt(diag(vcov(ml_lme_fit)))),
                       tau1 = tau[2],
                       tau2 = tau[1])
  
  
  # multilevel random-effects model using glmmTMB()
  if (method=="REML") REML <- TRUE else REML <- FALSE
  
  if (constant_var) {
    VCV <- diag(v_bar, nrow = nrow(dat))
    ml_glmm_fit <- glmmTMB(yi ~ 1 + equalto(0 + obs|g,VCV) + (1|district),
                            REML = REML,
                            data=dat)
    tau <- c(exp(ml_glmm_fit$fit$par)[["theta"]], sigma(ml_glmm_fit))
  } else {
    VCV <- diag(dat$vi)
    ml_glmm_fit <- glmmTMB(yi ~ 1 + equalto(0 + obs|g,VCV) + (1|district),
                           REML = REML,
                           data=dat)
    tau <- c(exp(ml_glmm_fit$fit$par)[["theta"]], sigma(ml_glmm_fit))
  }
  ml_glmm <- data.frame(f = "glmmTMB", 
                         logLik = logLik(ml_glmm_fit)[1],
                         df = attr(logLik(ml_glmm_fit), "df"), 
                         est = as.numeric(unlist(fixef(ml_glmm_fit))[[1]]), 
                         se = as.numeric(sqrt(vcov(ml_glmm_fit)[[1]])), 
                         tau1 = tau[1],
                         tau2 = tau[2])
  
  
  kable(rbind(ml_rma, ml_glmm, ml_lme))
  
}

Konstantopoulos("REML", constant_var = FALSE)
Konstantopoulos("REML", constant_var = TRUE)

Konstantopoulos("ML", constant_var = FALSE)
Konstantopoulos("ML", constant_var = TRUE)

```


A few things to note for this multilevel model :


-   The log likelihood is different for glmmTMB when using `REML` (but it is the same as metafor when using`ML`).
-   The standard error is different for glmmTMB to the fourth decimal point for all models (still quite close but not exactly the same). 





# Speed comparison

First let's test the speed of the simulated dataset with N=500 effect sizes across 50 studies.

```{r}
#| label: bench.fit
#| cache: TRUE

# for propto
colnames(VCV) <- 1:nrow(dat)
row.names(VCV) <- 1:nrow(dat)

# using the data from the second simulation - this runs 100 times by default
bench_fit <- microbenchmark(
  metafor = rma.mv(y, VCV, random = list(~1 | study, ~1 | id), 
                   test = "t",
                   dfs = "contain",
                   data=dat),
  metafor.sparse = rma.mv(y, VCV, random = list(~1 | study, ~1 | id), 
                   test = "t",
                   dfs = "contain",
                   sparse = TRUE, ##to check if it speeds up run time
                   data=dat),                 
  equalto = glmmTMB(y ~ 1 + (1|study) + equalto(0 + obs|g,VCV), 
                    data=dat,
                    REML=T),
  propto = glmmTMB(y ~ 1 + (1|study) + propto(0 + obs|g,VCV), # expect it to be the same
                    data=dat,
                    REML=T),
 equalto.sparse = glmmTMB(y ~ 1 + (1|study) + equalto(0 + obs|g,VCV), ### look into sparseX?
                    data=dat,
                    sparseX = c(cond=TRUE),
                    REML=T)
)

bench_fit
plot(bench_fit)

```

The glmmTMB implementations (equalto and propto) are showing a faster model run time (in seconds) compare to metafor default settings, however they are slower when metafor is used with the `sparse=TRUE` argument that speeds up matrix calculations.

Let's check the profiling of the glmmTMB model for this dataset to see which parts/functions take the most time:

```{r}
# Run profiling on model fitted with an average size dataset
Rprof(tmp <- tempfile())
fit.equalto <- glmmTMB(y ~ 1 + (1|study) + equalto(0 + obs|g,VCV),
                       data=dat,
                       REML=T)
Rprof()
prof <- summaryRprof(tmp)
unlink(tmp)


#self.time: Time spent within the function itself, excluding time spent in functions it calls.
self_df <- as.data.frame(prof$by.self) |>
  tibble::rownames_to_column("Function") |>
  dplyr::arrange(desc(self.pct))

#total.time: Time spent in the function including all the time in called functions.
total_df <- as.data.frame(prof$by.total) |>
  tibble::rownames_to_column("Function") |>
  dplyr::arrange(desc(total.pct))

# top 10 by self time
kable(head(self_df, 10), caption = "Top 10 Functions by Self Time (%)")

# top 10 by total time
kable(head(total_df, 10), caption = "Top 10 Functions by Total Time (%)")
```






What happens if we increase the dataset size by 10-fold (N=5000), let's simulate a new dataframe (`dat.big`) with 200 studies.

```{r}
#| cache: TRUE

# set up indices
set.seed(123)
k.studies <- 200 # number of studies
k.per.study <- 25 # number of effect sizes per study
study <- rep(seq_len(k.studies), times=k.per.study) # study id 
k <- length(study) # total number of effect sizes
id <- seq_len(k) # id of between study effect sizes 
es.id <- unlist(lapply(k.per.study, seq_len)) # id of within study effect sizes


# fixed effect coeff true value (i.e. the overall mean)
b0 <- 0

# simulate sampling errors variances
vi <- rbeta(k, 2, 20)

# get VCV matrix assuming within-study correlation of rho=0.5 
VCV <- matrix(0, nrow = k, ncol = k) 
for (i in 2:k) {
  for (j in 1:i) {
    if (study[i] == study[j]) {
      VCV[i,j] <- 0.5 * sqrt(vi[i] * vi[j]) 
    }
  }
}
VCV[upper.tri(VCV)] <- t(VCV)[upper.tri(VCV)] # fill in upper diagonal
diag(VCV) <- vi # fill in diagonal


# random effects
sigma2.u <- 0.2
sigma2.m <- 0.3
u <- rnorm(k.studies, 0, sqrt(sigma2.u))[study]
m <- rnorm(k, 0, sqrt(sigma2.m))
#e <- rnorm(k, 0, sqrt(vi))[study] #without within-study correlation (diag VCV)
e <- mvrnorm(n = 1, mu = rep(0, length(vi)), Sigma = VCV) #excluding within-study correlation


# compute y
y <- b0 + u + m + e


# combine into dataframe
dat.big <- data.frame(y = y,
                  vi = vi,
                  study = study,
                  id = id,
                  es.id = es.id,
                  obs = as.factor(id), # unique ID for propto and equalto (really important it is a factor variable)
                  g = 1 # group ID for propto and equalto
)
```

Run benchmark speed comparison again with `dat.big` for 10 runs (as it takes very long) and plot the results:

```{r}
#| label: bench.fit.big
#| cache: TRUE
#| eval: TRUE

# for propto
colnames(VCV) <- 1:nrow(dat.big)
row.names(VCV) <- 1:nrow(dat.big)

# using the data from the second simulation - this runs 100 times by default
bench_bigfit <- microbenchmark(
  metafor = rma.mv(y, VCV, random = list(~1 | study, ~1 | id),
                   test = "t",
                   dfs = "contain",
                   data=dat.big),
 metafor.sparse = rma.mv(y, VCV, random = list(~1 | study, ~1 | id), 
                   test = "t",
                   dfs = "contain",
                   sparse = TRUE, # to speed up run time
                   data=dat.big),   
  equalto = glmmTMB(y ~ 1 + (1|study) + equalto(0 + obs|g,VCV),
                    data=dat.big,
                    REML=T),
 times=5
)

bench_bigfit
plot(bench_bigfit)
```

Note: BLAS is installed on my session and seems to improve the speed performance of metafor more than glmmTMB. This results is over n=10 runs/times, there might be some variability in speed that is not captured here (I am planning to set up a simulation study to test this further).


Let's check the profiling of the glmmTMB model for this dataset to see which parts/functions take the most time:

```{r}
#| cache: TRUE

# Run profiling on model fitted with an average size dataset
Rprof(tmp <- tempfile())
fit.equalto <- glmmTMB(y ~ 1 + (1|study) + equalto(0 + obs|g,VCV),
                       data=dat.big,
                       REML=T)
Rprof()
prof <- summaryRprof(tmp)
unlink(tmp)


#self.time: Time spent within the function itself, excluding time spent in functions it calls.
self_df <- as.data.frame(prof$by.self) |>
  tibble::rownames_to_column("Function") |>
  dplyr::arrange(desc(self.pct))

#total.time: Time spent in the function including all the time in called functions.
total_df <- as.data.frame(prof$by.total) |>
  tibble::rownames_to_column("Function") |>
  dplyr::arrange(desc(total.pct))

# top 10 by self time
kable(head(self_df, 10), caption = "Top 10 Functions by Self Time (%)")

# top 10 by total time
kable(head(total_df, 10), caption = "Top 10 Functions by Total Time (%)")
```



# Next steps....

### Improve speed

Alternative strategy for `equalto` is to use Gaussian Markov Random Field in TMB cpp that uses the inverse of the covariance matrix: 
[density::GMRF_t](https://kaskr.github.io/adcomp/classdensity_1_1GMRF__t.html)

### Further notes

Notes for potential future improvements in implementation:

-   For `propto` and `equalto` do not show entire Std.Dev. and correlation matrix of the conditional model (only the `head()`?) in the summary output. When sample size is large it makes it unreadable.
-   For `propto` and `equalto` show the random effect component (not scaled).

Notes for future testing of equalto cov structure:

-   Make an example of a phylogenetic multilevel meta-analysis.
-   Make an example of a meta-analysis with a non-gaussian distribution (e.g. binomial, poisson) and compare it with `rma.glmm()` function in `metafor`. Speed here should be improved for glmmTMB. Maybe use this as a dataset: [https://www.metafor-project.org/doku.php/analyses:stijnen2010](https://www.metafor-project.org/doku.php/analyses:stijnen2010).


# Session info

```{r}
#| label: Reproducibility-SessionInfo-R-environment
#| fig-align: "center"
#| out-width: '100%'
#| results: asis
#| message: false
#| warnings: false

library(sessioninfo)
library(details)

si <- session_info()
details(si, summary = 'Current session info', open = FALSE)

```
