---
title: "Meta-analysis models in glmmTMB: pilot simulation study"
author: "Coralie Williams"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_width: 8
    fig_height: 6
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  #comment = "#>",
  fig.width=7.5,
  fig.height=5
)

options(rmarkdown.html_vignette.check_title = FALSE)

# load libraries
pacman::p_load(metafor,
               glmmTMB,
               ggplot2,
               tidyverse,
               cowplot,
               ggdark,
               ggdist,
               gridExtra,
               kableExtra,
               here, 
               ape)

```

This vignette describes the pilot simulation study results assessing the `equalto` covariance implementation in `glmmTMB` (Brooks et al., 2019). The simulation study aims to compare the new implementation with the `metafor` R package (Viechtbauer, 2010) for validation and proof-of-concept. 

For demonstration purposes, this pilot study assesses feasibility by simulating a single dataset, examining one condition per parameter, and fitting all models planned for the full simulation study.

We will investigate four effect size measures:

*   Standardised Mean Difference (SMD)

*   Log Response Ratio (lnRR) 

*   Odd ratio effect sizes (OR)

*   Poisson meta-analysis to model incidence rate ratios (IRR)


### Notation

- $y_i$: observed effect size for study $i$.
- $k$: number of studies.
- $\mu$: overall mean effect on the analysis scale (SMD, lnRR, log-OR, log-IRR).
- $\tau^2$: between-study (random-effects) variance.
- $\mathbf{V}$: block-diagonal sampling-error variance–covariance matrix.
- $\sigma_T, \sigma_C$: within-group standard deviations for treatment and control (SMD/lnRR).
- $\mu_C$: control-group mean on the original scale (SMD/lnRR).
- $n_T, n_C$: sample sizes in treatment and control groups (SMD/lnRR/OR).
- $t_T, t_C$: exposure times (person-time) for treatment and control groups (IRR).

\


### Model specification

For all effect size measures we will assess the random effect meta-analysis model:


$$y_{i} = \mu + u_{i} +  e_{i}  \tag{1}$$
$$u \sim N(0,\tau^2)$$
$$e \sim N(0,V)$$


Additionally, for non continuous effect size measures (OR and IRR), we will assess Binomial-Normal model and Poisson-Normal models.


### Pilot parameter settings


| Measure | $\mu$         | $\tau^2$ | $\sigma_T$ | $\sigma_C$ | $\mu_C$ | $n_T$ | $n_C$ | $t_T$ | $t_C$ |
|:--------|:--------------|:--------:|:----------:|:----------:|:-------:|:-----:|:-----:|:-----:|:-----:|
| SMD     | $0.3$         | $0.16$     | $10$       | $10$       | $100$   | $50$  | $50$  |  -    |  -    |
| lnRR    | $\log(1.3)$   | $0.16$     | $10$       | $10$       | $100$   | $50$  | $50$  |  -    |  -    |
| OR      | $\log(1.5)$   | $0.16$     |  -         |  -         |   -     | $50$  | $50$  |  -    |  -    |
| IRR     | $\log(1.5)$   | $0.16$     |  -         |  -         |   -     |  -    |  -    | $50$  | $50$  |



### Set up glmmTMB

<!-- # build package from branch: https://github.com/coraliewilliams/glmmTMB/tree/equalto_covstruc -->
<!-- # NOTE: remove glmmTMB (if installed) and reload R  -->
<!-- #remotes::install_github("coraliewilliams/glmmTMB", ref="equalto_covstruc", subdir="glmmTMB") -->


```{r}
# load glmmTMB from forked repo branch 
library(glmmTMB)

# check the available covstruc (you should see equalto in number 13)
(glmmTMB:::.valid_covstruct)

```






## Modelling standardised mean differences (SMD)

Load simulated dataset:

```{r}
# load simulated SMD data 
load(here("data", "dat_smd.RDATA"))
dat.smd <- out_smd
```


Fit random effects meta-analysis with metafor and glmmTMB:

```{r}
# fit metafor model
m.smd.rma <- rma(yi, vi, data=dat.smd)

# fit glmmTMB model
dat.smd$g <- 1
dat.smd$id <- as.factor(1:nrow(dat.smd))
V <- diag(dat.smd$vi, nrow = nrow(dat.smd))
m.smd.tmb <- glmmTMB(yi ~ 1 + equalto(0 + id|g,V), ## the map variable "id" must be a factor
                    REML = TRUE,
                    data=dat.smd)
```

Extract model estimates:

```{r, echo=FALSE}
# metafor output
m1.smd_rma <- data.frame(f = "rma.uni", 
                        logLik = logLik(m.smd.rma)[1],
                        df = attr(logLik(m.smd.rma), "df"), 
                        est = as.numeric(m.smd.rma$b), 
                        se = m.smd.rma$se, 
                        pvalue = m.smd.rma$pval,
                        tau2 = m.smd.rma$tau2
                        )

# glmmTMB output 
fix.tmb <- as.data.frame(summary(m.smd.tmb)$coefficients$cond)
m1.smd_tmb <- data.frame(f = "glmmTMB", 
                        logLik = logLik(m.smd.tmb)[1],
                        df = attr(logLik(m.smd.tmb), "df"), 
                        est = fix.tmb$Estimate, 
                        se = fix.tmb$`Std. Error`, 
                        pvalue = fix.tmb$`Pr(>|z|)`,
                        tau2 = sigma(m.smd.tmb)^2
                        )
  
kable(rbind(m1.smd_rma, m1.smd_tmb))
```


## Modelling log response ratios (lnRR)

Load simulated dataset:

```{r}
# load simulated lnRR data 
load(here("data", "dat_lnrr.RDATA"))
dat.lnrr <- out_lnrr
```


Fit random effects meta-analysis with metafor and glmmTMB:

```{r}
# fit metafor model
m.lnrr.rma <- rma(yi, vi, data=dat.lnrr)

# fit glmmTMB model
dat.lnrr$g <- 1
dat.lnrr$id <- as.factor(1:nrow(dat.lnrr))
V <- diag(dat.lnrr$vi, nrow = nrow(dat.lnrr))
m.lnrr.tmb <- glmmTMB(yi ~ 1 + equalto(0 + id|g,V), ## the map variable "id" must be a factor
                    REML = TRUE,
                    data=dat.lnrr)
```

Extract model estimates:

```{r}
# metafor output
m1.lnrr_rma <- data.frame(f = "rma.uni", 
                        logLik = logLik(m.lnrr.rma)[1],
                        df = attr(logLik(m.lnrr.rma), "df"), 
                        est = as.numeric(m.lnrr.rma$b), 
                        se = m.lnrr.rma$se, 
                        pvalue = m.lnrr.rma$pval,
                        tau2 = m.lnrr.rma$tau2
                        )

# glmmTMB output 
fix.tmb <- as.data.frame(summary(m.lnrr.tmb)$coefficients$cond)
m1.lnrr_tmb <- data.frame(f = "glmmTMB", 
                        logLik = logLik(m.lnrr.tmb)[1],
                        df = attr(logLik(m.lnrr.tmb), "df"), 
                        est = fix.tmb$Estimate, 
                        se = fix.tmb$`Std. Error`, 
                        pvalue = fix.tmb$`Pr(>|z|)`,
                        tau2 = sigma(m.lnrr.tmb)^2
                        )
  
kable(rbind(m1.lnrr_rma, m1.lnrr_tmb))
```



## Modelling log odds ratios (OR)

Set up simulated dataset:

```{r}
# load simulated OR data (with between study heterogeneity tau^2 = 0.16)
load(here("data", "dat_OR.RDATA"))
dat.or <- out_or

# set up long data format
n <- c(dat.or$a+dat.or$b, dat.or$c+dat.or$d)
event <- c(dat.or$a, dat.or$c)
study <- rep(1:nrow(dat.or), each=2)
obs <- 1:length(n)
treat <- c(rep(1, nrow(dat.or)), rep(0, nrow(dat.or)))
control <- 1 - treat
treat12 <- treat - 0.5
dat.or_long <- data.frame(study, obs, n, event, treat, control, treat12)

```

### Normal-Normal model

Fit random effects meta-analysis with metafor and glmmTMB:

```{r}
# fit metafor model
m.or.rma <- rma(yi, vi, data=dat.or)


# fit glmmTMB model
dat.or$g <- 1
dat.or$id <- as.factor(1:nrow(dat.or))
V <- diag(dat.or$vi, nrow = nrow(dat.or))

m.or.tmb <- glmmTMB(yi ~ 1 + equalto(0 + id|g,V), ## the map variable "id" must be a factor
                    REML = TRUE,
                    data=dat.or)
```

Extract model estimates:

```{r}
# fit metafor model
res.or_rma <- data.frame(f = "rma.uni", 
                        logLik = logLik(m.or.rma)[1],
                        df = attr(logLik(m.or.rma), "df"), 
                        est = as.numeric(m.or.rma$b), 
                        se = m.or.rma$se, 
                        pvalue = m.or.rma$pval,
                        tau2 = m.or.rma$tau2
                        )

# glmmTMB output 
fix.tmb <- as.data.frame(summary(m.or.tmb)$coefficients$cond)
res.or_tmb <- data.frame(f = "glmmTMB", 
                        logLik = logLik(m.or.tmb)[1],
                        df = attr(logLik(m.or.tmb), "df"), 
                        est = fix.tmb$Estimate, 
                        se = fix.tmb$`Std. Error`, 
                        pvalue = fix.tmb$`Pr(>|z|)`,
                        tau2 = sigma(m.or.tmb)^2
                        )
  
kable(rbind(res.or_rma, res.or_tmb))
```




### Binomial-Normal model

Fit models with with metafor and glmmTMB:

```{r}
m2.or.rma <- rma.glmm(measure="OR", 
                      ai=a, bi=b,
                      ci=c, di=d,
                      data=dat.or,
                      model="UM.RS",
                      nAGQ=1, #specifies Laplace approximation (it'll do it automatically; can't do nAGQ>1)
                      drop00 = F)


m2.or.tmb <- glmmTMB(cbind(event,n-event) ~ factor(treat) + (treat12-1|study) + (1|study), 
                         family = binomial(link = "logit"), 
                         data=dat.or_long)


m3.or.tmb <- glmmTMB(cbind(event,n-event) ~ factor(treat) + (control+treat-1|study), 
                         family = binomial(link = "logit"), 
                         data=dat.or_long)


```


Get estimates to compare all models:


```{r}
res1.or_rma <- data.frame(f = "rma.uni", 
                        logLik = logLik(m.or.rma)[1],
                        df = attr(logLik(m.or.rma), "df"), 
                        est = as.numeric(m.or.rma$b), 
                        se = m.or.rma$se, 
                        pvalue = m.or.rma$pval,
                        tau2 = m.or.rma$tau2
                        )


res1.or_tmb <- data.frame(f = "glmmTMB", 
                        logLik = logLik(m.or.tmb)[1],
                        df = attr(logLik(m.or.tmb), "df"), 
                        est = as.numeric(unlist(fixef(m.or.tmb))[[1]]), 
                        se = as.numeric(sqrt(vcov(m.or.tmb)[[1]])), 
                        pvalue = summary(m.or.tmb)$coefficients$cond[1, "Pr(>|z|)"],
                        tau2 = sigma(m.or.tmb)^2
                        )

res2.or_rma <- data.frame(f = "rma.glmm", 
                        logLik = logLik(m2.or.rma)[1],
                        df = attr(logLik(m2.or.rma), "df"), 
                        est = as.numeric(m2.or.rma$b), 
                        se = m2.or.rma$se, 
                        pvalue = m2.or.rma$pval,
                        tau2 = m2.or.rma$tau2
                        )

res2.or_tmb <- data.frame(f = "glmmTMB.binomial.1", 
                        logLik = logLik(m2.or.tmb)[1],
                        df = attr(logLik(m2.or.tmb), "df"), 
                        est = as.numeric(unlist(fixef(m2.or.tmb))[2]), 
                        se = as.numeric(sqrt(vcov(m2.or.tmb)$cond[2,2])), 
                        pvalue = summary(m2.or.tmb)$coefficients$cond[2, "Pr(>|z|)"],
                        tau2 = c((VarCorr(m2.or.tmb)$cond$study))
                        )

res3.or_tmb <- data.frame(f = "glmmTMB.binomial.2", 
                        logLik = logLik(m3.or.tmb)[1],
                        df = attr(logLik(m3.or.tmb), "df"), 
                        est = as.numeric(unlist(fixef(m3.or.tmb))[2]), 
                        se = as.numeric(sqrt(vcov(m3.or.tmb)$cond[2,2])), 
                        pvalue = summary(m3.or.tmb)$coefficients$cond[2, "Pr(>|z|)"],
                        tau2 = VarCorr(m3.or.tmb)$cond$study[1,1]
                        +VarCorr(m3.or.tmb)$cond$study[2,2]
                        -2*VarCorr(m3.or.tmb)$cond$study[1,2]
                      )


kable(rbind(res1.or_rma, res1.or_tmb, res2.or_rma, res2.or_tmb, res3.or_tmb))


```


## Modelling log incidence ratios (IRR)


Load simulated dataset:

```{r}
# load simulated IRR data (with between study heterogeneity tau^2 = XXX)
load(here("data", "dat_IRR.RDATA"))
dat.irr <- out_irr
```

### Normal-normal model

Fit random effects meta-analysis with metafor and glmmTMB:

```{r}
# fit metafor model
m.irr.rma <- rma(yi, vi, data=dat.irr)

# fit glmmTMB model
dat.irr$g <- 1
dat.irr$id <- as.factor(1:nrow(dat.irr))
V <- diag(dat.irr$vi, nrow = nrow(dat.irr))
m.irr.tmb <- glmmTMB(yi ~ 1 + equalto(0 + id|g,V), ## the map variable "id" must be a factor
                    REML = TRUE,
                    data=dat.irr)
```


Extract estimates from model:

```{r}
# fit metafor model
m1.irr_rma <- data.frame(f = "rma.uni", 
                        logLik = logLik(m.irr.rma)[1],
                        df = attr(logLik(m.irr.rma), "df"), 
                        est = as.numeric(m.irr.rma$b), 
                        se = m.irr.rma$se, 
                        pvalue = m.irr.rma$pval,
                        tau2 = m.irr.rma$tau2
                        )

# glmmTMB output 
fix.tmb <- as.data.frame(summary(m.irr.tmb)$coefficients$cond)
m1.irr_tmb <- data.frame(f = "glmmTMB", 
                        logLik = logLik(m.irr.tmb)[1],
                        df = attr(logLik(m.irr.tmb), "df"), 
                        est = fix.tmb$Estimate, 
                        se = fix.tmb$`Std. Error`, 
                        pvalue = fix.tmb$`Pr(>|z|)`,
                        tau2 = sigma(m.irr.tmb)^2
                        )
  
kable(rbind(m1.irr_rma, m1.irr_tmb))
```



### Poisson-Normal model

```{r}
# set up long data format

```

Fitting with metafor:

```{r}
#rma.glmm(measure="IRR", x1i=x1i, t1i=t1i, x2i=x2i, t2i=t2i, data=dat, model="CM.EL")
```


## Extra - example case studies

### Bivariate meta-analysis

Here I use the dataset of meta-analysis on the effectiveness of the BCG vaccine against tuberculosis (Colditz et al., 1994), which is available within the `metafor` package. 

```{r}
data(dat.bcg)
dat_long <- to.long(measure="OR", ai=tpos, bi=tneg, ci=cpos, di=cneg, data=dat.bcg)
levels(dat_long$group) <- c("exp", "con")
dat_long$group <- relevel(dat_long$group, ref="con")
dat_long <- escalc(measure="PLO", xi=out1, mi=out2, data=dat_long)
dat_long$obs <- as.factor(1:nrow(dat_long))
dat_long$g <- 1
```

Fit bivariate meta-analysis with metafor and glmmTMB:

```{r}
# fit with metafor
bv_rma <- rma.mv(yi, vi, mods = ~ group, 
                 random = ~ group | study, 
                 struct = "UN",
                 method = "REML",
                 data=dat_long)

# fit with glmmTMB
VCV <- diag(dat_long$vi) #variance covariance matrix for sampling errors
bv_TMB <- glmmTMB(yi ~ group + (0 + group|study) + equalto(0 + obs|g,VCV),
                  dispformula = ~ 0, #set residual variance to zero
                  REML = "TRUE",
                  data=dat_long)

```

Get models output:

```{r}

# metafor output
res.bv_rma <- with(bv_rma, data.frame(f = "rma.mv",
                                      logLik = logLik(bv_rma),
                                      df = attr(logLik(bv_rma), "df"),
                                      est.b0 = as.numeric(b[1]),
                                      se.b0 = se[1],
                                      est.b1 = as.numeric(b[2]),
                                      se.b1 = se[2],
                                      sigma2.con = tau2[1],
                                      sigma2.exp = tau2[2],
                                      rho = rho))

# glmmTMB output
vc <- VarCorr(bv_TMB) 
var_con <- vc$cond$study["groupcon", "groupcon"]
var_exp <- vc$cond$study["groupexp", "groupexp"]
rho.tmb <- vc$cond$study["groupcon", "groupexp"] / sqrt(var_con * var_exp)

res.bv_tmb <- data.frame(f = "glmmTMB", 
                      logLik = logLik(bv_TMB)[1],
                      df = attr(logLik(bv_TMB), "df"),                         
                      est.b0 = as.numeric(unlist(fixef(bv_TMB))[[1]]), 
                      se.b0 = as.numeric(sqrt(vcov(bv_TMB)[[1]]))[1], 
                      est.b1 = as.numeric(unlist(fixef(bv_TMB))[[2]]), 
                      se.b1 = as.numeric(sqrt(vcov(bv_TMB)[[1]]))[4], 
                      sigma2.con = var_con, 
                      sigma2.exp = var_exp, 
                      rho = rho.tmb)

  
kable(rbind(res.bv_rma, res.bv_tmb))

```



### Phylogenetic meta-analysis with dependent effect sizes

Here I demonstrate how to fit a phylogenetic meta-analysis with dependent effect sizes with glmmTMB using a simulated dataset.


$$y_{ijk} = \mu + u_{ij} + s_i + n_k + p_k + e_{ij}$$ 
$$u \sim N(0,\sigma^2_{u}I_u)$$ 
$$s \sim N(0, \sigma^2_sI_s)$$ 
$$n \sim N(0, \sigma^2_nI_n)$$ 
$$p \sim N(0, \sigma^2_p\textbf{A})$$ 
$$e \sim N(0,V)$$

where $\sigma^2_s$ the study level variance component, $\sigma^2_u$ the estimate level variance component, 
$\sigma^2_n$ is the variance of the species level random effect, $\sigma^2_p$ is the variance of phylogenetic random effect, $\textbf{V*}$ the block diagonal variance-covariance matrix of the sampling error with effect sizes from the same study, and $\textbf{A}$ is the phylogenetic correlation matrix based on a Brownian model of evolution.



```{r}
# Load some phylo dataset and matrix
load(here("data", "phylo_simulated_data.RData"))
```


Fit model using metafor:

```{r}
phylo_rma_fit <- rma.mv(yi,
                   V,
                   random = list(~ 1| id, ~ 1 | study, ~ 1 | species, ~ 1 | species.phylo),
                   R = list(species.phylo=P),
                   test = "t",
                   dfs = "contain",
                   data = dat)
```


Fit model using glmmTMB:

```{r}
phylo_tmb_fit <- glmmTMB(yi ~ 1 + equalto(0 + id|g,V) + (1|study) + (1|species) + propto(0 + species|g,P),
                       data=dat,
                       REML=T)
```


Compare output of metafor and glmmTMB:

```{r}
# metafor output
phylo_rma <- with(phylo_rma_fit,
                  data.frame(f = "rma.mv", 
                        logLik = logLik(phylo_rma_fit),
                        df = attr(logLik(phylo_rma_fit), "df"), 
                        est = as.numeric(b), 
                        se = se, 
                        sigma.u = sqrt(sigma2[1]), 
                        sigma.s = sqrt(sigma2[2]),
                        sigma.n = sqrt(sigma2[3]),
                        sigma.p = sqrt(sigma2[4])
                        ))

# glmmTMB output
re_tmb <- as.data.frame(confint(phylo_tmb_fit, parm = "theta_"))
fix_tmb <- as.data.frame(summary(phylo_tmb_fit)$coefficients$cond)

phylo_tmb <- data.frame(f = "glmmTMB", 
                        logLik = logLik(phylo_tmb_fit )[1],
                        df = attr(logLik(phylo_tmb_fit), "df"), 
                        est = fix_tmb$Estimate, 
                        se = fix_tmb$`Std. Error`, 
                        sigma.u = sigma(phylo_tmb_fit), 
                        sigma.s = re_tmb$Estimate[1],
                        sigma.n = re_tmb$Estimate[2],
                        sigma.p = re_tmb$Estimate[3] 
                        )
  
  
kable(rbind(phylo_rma, phylo_tmb))
```






## References

    Brooks, M. E., Kristensen, K., van Benthem, K. J., Magnusson, A., Berg, C. W., Nielsen, A., Skaug, H. J., Mächler, M., & Bolker, B. M. (2017). glmmTMB balances speed and flexibility among packages for zero-inflated generalized linear mixed modeling. R Journal, 9(2), 378–400. https://doi.org/10.32614/RJ-2017-066

    Colditz, G. A., Brewer, T. F., Berkey, C. S., Wilson, M. E., Burdick, E., Fineberg, H. V., & Mosteller, F. (1994). Efficacy of BCG Vaccine in the Prevention of Tuberculosis: Meta-analysis of the Published Literature. JAMA, 271(9), 698–702. https://doi.org/10.1001/jama.1994.03510330076038

    Jackson, D., Law, M., Stijnen, T., Viechtbauer, W., & White, I. R. (2018). A comparison of seven random-effects models for meta-analyses that estimate the summary odds ratio. Statistics in Medicine, 37(7), 1059–1085. https://doi.org/10.1002/sim.7588

    Viechtbauer, W. (2010). Conducting meta-analyses in R with the metafor package. Journal of Statistical Software, 36(3), 1-48. https://cran.r-project.org/web/packages/metafor/index.html


