---
title: "Fitting meta-analysis with `glmmTMB`"
author: "Coralie Williams"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    theme: cosmo
    smooth-scroll: true
    toc-title: "**Table of contents**"
    embed-resources: true
    code-fold: show
    code-tools: true
    number-sections: true
    number-depth: 2
    colorlinks: true
    footnote-location: margin
---

This webpage is supporting material to the paper "*Meta-analysis with `glmmTMB` R package*".

We provide show how to fit a meta-analysis using the `equalto` covariance structure in `glmmTMB` R package, which allows to specify the sampling error variance-covariance (VCV) matrix.

### Contact {#contact}

For questions or comments, contact Coralie Williams: [coraliewilliams\@outlook.com](mailto:coraliewilliams@outlook.com)


# Introduction 

To demonstrate how to fit a meta-analysis using the `glmmTMB` R package, we will fit a multilevel meta-analytic model assuming multiple effect sizes per study. This model synthesises effect sizes across studies by accounting for variation at both the study and observation (effect size) levels, while also modelling correlated sampling errors within studies.

Given $j$ effect sizes within $i$ studies, we specify the model as:

$$
y_{ij} = b_0 + u_i + m_{ij} + e_{ij},
$$

where $b_0$ is the overall mean effect, $u_i \sim N(0, \sigma^2_{u}I)$ represents the among-study random effect, $m_{ij} \sim N(0, \sigma^2_{m}I)$ captures the within-study random effect (at the observation-level), and $e_{ij} \sim N(0, \text{VCV})$ are the sampling errors at the observation-level.

$\text{VCV}$ represents a block-diagonal variance-covariance matrix including known (or calculated) sampling variances $\sigma^2_{e}=v_i$ on the diagonal and accounts for correlations among sampling errors from the same study.

### Load `glmmTMB` package

Currently, `equalto` is not on CRAN (yet!), so you need to install `glmmTMB` from the [following branch](https://github.com/coraliewilliams/glmmTMB/tree/equalto_covstruc):

::: callout-note
**Important note:** If you have `glmmTMB` already installed, you will need to uninstall it first.
:::

```{r}
#| eval: false
#| output: false
#| warnings: false

# build package from branch: 
remotes::install_github("coraliewilliams/glmmTMB", ref="equalto_covstruc", subdir="glmmTMB")
```

Load `glmmTMB` package:

```{r}
library(glmmTMB)
```

Check if the `equalto` covariance structure is available (it should be in position 13 of this output) :

```{r}
glmmTMB:::.valid_covstruct
```

### Simulate some data

Now, let's simulate some data following the structure of the above multilevel meta-analysis.

```{r}
#| output: false
#| warnings: false

# load libraries
suppressPackageStartupMessages({
  library(MASS)
  library(kableExtra)
  library(metafor)
})

# set up indices
k.studies <- 10 # number of studies
k.per.study <- 3 # number of effect sizes per study
study <- rep(seq_len(k.studies), times=k.per.study) # study id 
k <- length(study) # total number of effect sizes
id <- seq_len(k) # id of between study effect sizes 
es.id <- unlist(lapply(k.per.study, seq_len)) # id of within study effect sizes


# fixed effect coeff true value (i.e. the overall mean)
b0 <- 0

# simulate sampling errors variances
set.seed(123)
vi <- rbeta(k, 2, 20)

# get VCV matrix assuming within-study correlation of rho=0.5 
V <- matrix(0, nrow = k, ncol = k) 
for (i in 2:k) {
  for (j in 1:i) {
    if (study[i] == study[j]) {
      V[i,j] <- 0.5 * sqrt(vi[i] * vi[j]) 
    }
  }
}
V[upper.tri(V)] <- t(V)[upper.tri(V)] # fill in upper diagonal
diag(V) <- vi # fill in diagonal


# random effects
sigma2.u <- 0.2
sigma2.m <- 0.3
u <- rnorm(k.studies, 0, sqrt(sigma2.u))[study]
m <- rnorm(k, 0, sqrt(sigma2.m))

# simulate multivariate normal sampling errors with specific VCV
set.seed(123)
e <- mvrnorm(n = 1, mu = rep(0, length(vi)), Sigma = V) 
#e <- rnorm(k, 0, sqrt(vi))[study] #without within-study correlation (diag VCV)

# compute y
y <- b0 + u + m + e


# combine into dataframe
dat <- data.frame(y = y,
                  vi = vi,
                  study = study,
                  id = id,
                  es.id = es.id,
                  obs = as.factor(id), #unique ID for propto and equalto (really important it is a factor variable)
                  g = 1 #group ID for propto and equalto (a constant here)
)
```

### Fit `glmmTMB` model

Before we fit the model we need to obtain the within-study error variance-covariance matrix (which accounts for within-study correlation) assuming a constant correlation of $\rho=0.5$. The correlation between effect sizes from different studies is assumed to be zero.

We use the metafor `vcalc` function to obtain a block diagonal VCV matrix for the sampling errors:

```{r}
VCV <- metafor::vcalc(vi, cluster=study, obs=obs, data=dat, rho=0.5)
```

Now we fit the `glmmTMB` model with `equalto` covariance structure.

We specify:

-   the study-level random effect $u_i$ as `(1|study)`.

-   the sampling errors $e_{ij}$ are specified using the covariance structure `equalto(0 + obs|g,VCV)` allowing to specify the VCV matrix of the sampling errors.

-   the within-study random effect $m_{ij}$ that captures the variability at the observation level is modelled as the *residual error* in this model.

::: callout-note
**Important note:** the `equalto` multivariate random effect variable (here it is called `obs`) must be a factor and the corresponding VCV matrix must have rows and columns matching the levels of the `obs` variable.
:::


```{r}

# Fit glmmTMB model with equalto ----------------------------------
fit.equalto <- glmmTMB(y ~ 1 + (1|study) + (1|study) + equalto(0 + obs|g,VCV),
                       data=dat,
                       REML=T)

summary(fit.equalto)
```

### Fit `metafor` model

Let's compare with the `metafor` package `rma.mv` function that fits a multilevel model with random effects. Here the model is specified with two random effects: `(1|study)` corresponding to the among-study ($u_i$) and `(1|id)` corresponding to the within-study ($m_{ij}$) random effects. The sampling errors $e_{ij}$ VCV matrix is specified using the `V` argument.

```{r}
# Fit metafor model --------------------------------------------
fit.metafor <- rma.mv(yi = y, 
                      V = VCV,
                      random = list(~1 | study, ~1 | id),
                      control=list(REMLf=FALSE), #set this to obtain the exact same likelihood as glmmTMB 
                      data = dat) #uses REML by default

summary(fit.metafor)
```

Then, we extract the model results for each model and compare outputs:

```{r}
### Compare output 
equalto <- data.frame(model = "glmmTMB",
                      est = summary(fit.equalto)$coefficients$cond[,"Estimate"],
                      se = summary(fit.equalto)$coefficients$cond[,"Std. Error"],
                      zval = summary(fit.equalto)$coefficients$cond[,"z value"],
                      pval = summary(fit.equalto)$coefficients$cond[,"Pr(>|z|)"],
                      sigma.u = VarCorr(fit.equalto)$cond$study[,"(Intercept)"], ##among study variance estimate
                      sigma.m = sigma(fit.equalto)^2) ##within study variance estimate


metafor <- data.frame(model = "metafor", 
                      est = fit.metafor$b[[1]], 
                      se = fit.metafor$se[[1]], 
                      zval = fit.metafor$zval,
                      pval = fit.metafor$pval,
                      sigma.u = fit.metafor$sigma2[1], ## among study variance
                      sigma.m = fit.metafor$sigma2[2]) ## within study variance


output <- rbind(equalto, metafor)
kable(output)

```

The overall mean estimate and random effect variance estimates are identical across the two packages ✅


### Extra analyses

When the number of studies is small (<10) it is preferable to use the $t$-distribution and adjusted degrees of freedom to avoid Type I error rate of coefficient tests and low coverage rates of coefficient confidence intervals ([read more about the recommendations here](https://wviechtb.github.io/metafor/reference/misc-recs.html)).

In `metafor` you can do this directly in the `rma.mv` function when fitting a multilevel/multivariate model with the arguments `test="t"` and `dfs="contain"` (actually if `dfs="contain"` is set this automatically sets `test="t"`).

```{r}
# Fit metafor model --------------------------------------------
fit.metafor.t <- rma.mv(yi = y, 
                      V = VCV,
                      random = list(~1 | study, ~1 | id),
                      test = "t",
                      dfs = "contain", 
                      data = dat) 

summary(fit.metafor.t)
```

With `glmmTMB` you can derive the $t$-test statistic, adjusted degrees of freedom and corresponding P value the following way:

```{r}
# summary stats of fixed effects
fef <- as.data.frame(summary(fit.equalto)$coefficients$cond)
b0.est <- fef$Estimate
b0.se <- fef$`Std. Error`

# if the coefficient of interest varies at the level of corresponding to a random effect (i.e. here it is `study`) then the degree of freedom if the number of unique levels minus the number of estimated coefficients in the model
df.contain <- length(unique(dat$study)) - nrow(fef)
# otherwise it is the total nuber of estimates in the analysis minus the total number of model coefficients
#df.adj <- nrow(dat) - nrow(fef)

# get t-test statistic and p-value
t <- b0.est / b0.se
p <- 2 * pt(abs(t), df = df.contain, lower.tail = FALSE) # two-sided p value
# 95% CI
crit <- qt(0.975, df = df.contain)
ci   <- c(b0.est - crit * b0.se, b0.est + crit * b0.se)

# glmmTMB t-test output
fef.t <- data.frame(Estimate = b0.est,
                    SE = b0.se,
                    tval = t,
                    df = df.contain,
                    pval = p,
                    ci.lb = ci[1],
                    ci.ub = ci[2])
fef.t

```



# Illustrative examples

To illustrate how to fit meta-analyses with `glmmTMB`, we re-analyse three published datasets from distinct research domains:

- **Medicine** — vaccine effectiveness.
- **Ecology & evolution** — plant diversity effects.
- **Social sciences** — school-based learning programs.

Across these case studies, we show model specification, output, and interpretation using `glmmTMB`.


## Bivariate meta-analysis

Meta-analysis is widely used in medicine to combine evidence across clinical trials and observational studies. As many outcomes are events in medicine, common effect sizes measures used are odds ratios, risk ratios, risk differences, and their log transforms.

We will use the Colditz et al. (1994) dataset that is directly avaiable in `metafor` via the `metadat` R package. This published meta-analysis assessed the effectiveness of the Bacillus Calmette-Guerin (BCG) vaccine against tuberculosis across 13 studies. We will fit a bivariate meta-analysis that treats the two vaccination groups as separate outcomes, and estimates, for each group, the overall mean effect and among-study variance, as well as the correlation of study-level effects between groups (i.e., how strongly they co-vary).

```{r}
dat <- dat.colditz1994
```

Reshape to long format so that each arm has its own row (“v” = vaccinated, “c” = unvaccinated/control), and setting control as the reference group.

```{r}
dat_long <- to.long(measure="OR", ai=tpos, bi=tneg, ci=cpos, di=cneg, data=dat.bcg)
levels(dat_long$group) <- c("c", "v")
dat_long$group <- relevel(dat_long$group, ref="c")
```

We compute the logit-transformed event proportion (`PLO`) as the effect-size measure. And we also need to create a constant grouping dummy variable `g` for `glmmTMB` model specification (this needs to be a constant value for all rows). 

```{r}
dat_long <- escalc(measure="PLO", xi=out1, mi=out2, data=dat_long)
dat_long$obs <- as.factor(1:nrow(dat_long))
dat_long$g <- 1 #dummy variable
head(dat_long)
```

Derive the variance-covariance matrix (VCV) with in diagonal the sampling error variances and zero on the off-diagonals (assume studies are independent):

```{r}
VCV <- diag(dat_long$vi)
VCV[1:5,1:5] #show first five measures of VCV matrix
```

To fit a bivariate meta-analysis with `glmmTMB` using `equalto` supply the sampling-error variance–covariance matrix (here it is called `VCV`), and fix the dispersion at zero (`dispformula = ~ 0`) since we don't want to model any within-study variance effect. 

```{r}
bv_fit <- glmmTMB(yi ~ group + (0 + group|study) + equalto(0 + obs|g,VCV),
                       dispformula = ~ 0,
                       REML = TRUE,
                       data=dat_long)
```

To obtain the variance component estimates and corresponding correlation $\rho$ (i.e. `rho`) for each group we use the `VarCorr` function:

```{r}
vc <- VarCorr(bv_fit)
var_con <- vc$cond$study["groupc", "groupc"]
var_exp <- vc$cond$study["groupv", "groupv"]
rho.tmb <- vc$cond$study["groupc", "groupv"] / sqrt(var_con * var_exp)
```

Then we can extract the model results as follows: 

```{r}
bv_out <- data.frame(est.b0 = as.numeric(unlist(fixef(bv_fit))[[1]]), 
                     se.b0 = as.numeric(sqrt(vcov(bv_fit)[[1]]))[1], 
                     est.b1 = as.numeric(unlist(fixef(bv_fit))[[2]]), 
                     se.b1 = as.numeric(sqrt(vcov(bv_fit)[[1]]))[4], 
                     tau2_1 = var_con, #study-level heterogeneity for group=control
                     tau2_2 = var_exp, #study-level heterogeneity for group=exp
                     rho = rho.tmb)

bv_out
```



This is equivalent to the following model in `metafor`:

```{r}
bv_fit_rma <- rma.mv(yi, vi, mods = ~ group, 
                     random = ~ group | study, 
                     struct = "UN",
                     data=dat_long)
```

We can extract the model results as follows:

```{r}
bv_out_rma <- with(bv_fit_rma, data.frame(est.b0 = as.numeric(b[1]),
                                      se.b0 = se[1],
                                      est.b1 = as.numeric(b[2]),
                                      se.b1 = se[2],
                                      tau2_1 = tau2[1],
                                      tau2_2 = tau2[2],
                                      rho = rho))
bv_out_rma
```




## Phylogenetic meta-analysis

In meta-analysis of ecological and evolutionary studies, effect sizes are often drawn from multiple species. Because species share evolutionary history, these effect size estimates are not independent. To account for this phylogenetic relatedness, we can include a species-level random effect structured by a phylogenetic correlation (which is derived from a phylogenetic tree and assumed model of evolution). We can fit such a random effect in `glmmTMB` with the `propto` covariance structure. 

We will use a published meta-analysis dataset of plant diversity effects on leaf traits (Felix et al., 2023) to showcase how to use `glmmTMB` to fit a phylogenetic meta-analysis.

```{r}
#| results: "hide"    
#| message: false     
#| warning: false     

library(readr)
ef <- read_csv("felix_data/effect_sizes.csv")
options(na.action = "na.pass")
```

Data set-up:

```{r}
#| results: "hide"      # hide printed output
#| message: false      # hide messages
#| warning: false      # hide warnings

# load libraries
suppressPackageStartupMessages({
  library(rotl)
  library(ape)
  library(dplyr)
})

#remove empty rows in Nfixing
ef <- ef[which(ef$Nfixing!=is.na(ef$Nfixing)),] 

# construct phylogentic tree matrix for use as random factor
species <- unique(ef$species) # list of unique species in meta-analysis
species <- as.character(species) # change to character object
taxa <- tnrs_match_names(species)
tree <- rotl::tol_induced_subtree(taxa$ott_id)
tree$tip.label <- strip_ott_ids(tree$tip.label, remove_underscores = TRUE) # change ids to the names from dataset

# calculate correlations between all species = cor
tree2 <- compute.brlen(tree)
phylo.cor <- vcv(tree2, cor = T)

# add phylo random factor
ef$phylo <- factor(ef$species, levels = rownames(phylo.cor))

# set up phylo.cor row and column names
phylo.cor <- phylo.cor[rownames(phylo.cor) %in% levels(ef$phylo),
                       colnames(phylo.cor) %in% levels(ef$phylo)]

# add g variable
ef$g <- 1

# derive effect size level variable => make sure it is a factor
ef$id <- as.factor(1:nrow(ef))

# set up sampling errors variance-covariance matrix 
# assuming within-study a constant correlation of 0.5
V <- metafor::vcalc(vi, cluster=ACC, obs=id, data=ef, rho=0.5)

# make sure row matrix and id levels match
rownames(V) <- colnames(V) <- levels(ef$id)

```

Let's view the five rows of the species correlation matrix `phylo.cor`  and the sampling error variance-covariance matrix `V` (which assumes no within-study or among effect size correlations):

```{r}
phylo.cor[1:5,1:5]
V[1:5,1:5]
```

We fit a multilevel model using experimental site (\texttt{experiment}), study ID (\texttt{ACC}), individual effect ID (\texttt{id}) and plant species (\texttt{species}) were included as random factors to account for non-independence among effect sizes. 

This multilevel phylogenetic meta-analysis has two covariance structures in `glmmTMB`:

-   `propto`: to model shared evolutionary history across species assuming a proportional phylogenetic correlation matrix
-   `equalto`: to incorporate sampling-errors variance-covariance 


```{r}
#| cache: true
fit_phylo <- glmmTMB(yi ~ Nfixing-1 + 
                       (1 | experiment) + (1 | ACC) +
                       (1 | species) + propto(0 + phylo|g, phylo.cor) + 
                       equalto(0 + id|g, V), 
                     data = ef, 
                     REML=TRUE)

```

We can extract and view the fixed effect estimates, showing the two groups for `Nfixing` (whether neighboring plants have nitrogen fixing or not):

```{r}
summary(fit_phylo)$coefficients$cond
```

The variance component estimates can be obtained:

```{r}
# get random-effects varcor parameters
re_est <- as.data.frame(confint(fit_phylo, parm="theta_"))

# summarise random-effects variance estimates
data.frame(sigma2.experiment = round(re_est$Estimate[1]^2,4), 
           sigma2.ACC = round(re_est$Estimate[2]^2,4),
           sigma2.species = round(re_est$Estimate[3]^2,4),
           sigma2.phylo = round(re_est$Estimate[4]^2,4), 
           sigma2.id=round(sigma(fit_phylo)^2,4))

```



This model is equivalent to the following model fitted with `metafor`:

```{r}
#| cache: true
fit_phylo_rma <- rma.mv(yi, vi,
                        mods = ~ Nfixing-1,
                        random = list( ~ 1 | experiment, ~ 1 | ACC,
                                       ~1 | species, ~1 | phylo,
                                       ~1 | id), 
                        R = list(phylo = phylo.cor),
                        data = ef,
                        sparse = TRUE,
                        method="REML") #default is REML

summary(fit_phylo_rma)
```

We can store time and compare with `metafor` with and without using the argument `sparse=TRUE` which helps speed up computational speed:

```{r}
#| cache: true
## Fit glmmTMB model and store runtime
time_glmmTMB <- system.time(
  fit_phylo <- glmmTMB(
    yi ~ Nfixing - 1 +
      (1 | experiment) + (1 | ACC) +
      (1 | species) +
      propto(0 + phylo | g, phylo.cor) +
      equalto(0 + id | g, V),
    data = ef,
    REML = TRUE
  )
)[["elapsed"]]

## Fit rma.mv model and store runtime
time_rma <- system.time(
  fit_phylo_rma <- rma.mv(
    yi, vi,
    mods   = ~ Nfixing - 1,
    random = list(
      ~ 1 | experiment,
      ~ 1 | ACC,
      ~ 1 | species,
      ~ 1 | phylo,
      ~ 1 | id
    ),
    R      = list(phylo = phylo.cor),
    data   = ef,
    method = "REML"
  )
)[["elapsed"]]

## Fit rma.mv model and store runtime
time_rma_sparse <- system.time(
  fit_phylo_rma <- rma.mv(
    yi, vi,
    mods   = ~ Nfixing - 1,
    random = list(
      ~ 1 | experiment,
      ~ 1 | ACC,
      ~ 1 | species,
      ~ 1 | phylo,
      ~ 1 | id
    ),
    R      = list(phylo = phylo.cor),
    data   = ef,
    sparse = TRUE,
    method = "REML"
  )
)[["elapsed"]]

## Put runtimes into a tidy object
runtime_df <- data.frame(
  model       = c("glmmTMB_phylo", "rma.mv_phylo", "rma.mv_phylo_sparse"),
  runtime_sec = c(time_glmmTMB, time_rma, time_rma_sparse)
)

runtime_df
```




## Location-scale meta-analysis

In a standard random-effects meta-analysis, the between-study variance $\tau^2$ is assumed constant across studies. This is often unrealistic. **Location–scale models** relax this assumption by modelling both the *location* (the mean effect) and the *scale* (the amount of heterogeneity) as functions of study-level moderators. In practice, you can include different predictors in the mean and variance parts to test whether heterogeneity varies with study features (e.g., outcome type, design) or differs across subgroups. This yields a more flexible meta-analytic model that can explain, not just accommodate, between-study variability.

We will use a social-science meta-analysis of 48 studies on school-based writing-to-learn interventions. Each study contrasts an intervention group (greater emphasis on writing tasks) with a control group. The effect size measures used in this dataset is the *standardised mean differences*, where positive values indicate better academic performance under the intervention.

```{r}
dat <- dat.bangertdrowns2004
dat$id <- as.factor(dat$id)
dat$g <- rep(1, nrow(dat))
```

Define `V` - the sampling error variance-covariance matrix assuming a within-study correlation of 0.6:

```{r}
V <- diag(dat$vi)
row.names(V) <- colnames(V) <- levels(dat$id)
```

We fit a simple location-scale meta-analysis modelling the study sample size variable (`ni`) in the location and scale parts. We use `dispformula` in `glmmTMB` to specify the scale part of the model. 

```{r}
#| results: "hide"    
#| message: false     
#| warning: false   
fit_ls <- glmmTMB(yi ~ ni + equalto(0 + id|g, V),
                  dispformula = ~ ni,
                  data = dat,
                  REML = TRUE)
```

View the estimates of the location-part of the model:

```{r}
summary(fit_ls)$coefficients$cond
```

View the estimates of the scale-part of the model:

```{r}
summary(fit_ls)$coefficients$disp
```


We can fit the same model with `metafor` with the `rma` function and using the `scale` argument:

```{r}
fit_ls_rma <- rma(yi, vi, 
                  mods = ~ ni,
                  scale = ~ ni,
                  control=list(REMLf=FALSE), #this is to get the same log-likelihood
                  data = dat)
```


By default `metafor` uses the variance for the scale, whereas `glmmTMB` uses the standard deviation. Hence, to compare both outputs we need to times by x2 the estimates from the `glmmTMB` model (given the scale model uses a log link).


# Useful notes and links

## Changing optimization method

We can change the optimization method using by modifying the `control = glmmTMBControl()` argument. By default, `glmmTMB` uses the `nlminb` non-linear optimizer for parameter estimation. 



-   [Troubleshooting with glmmTMB](https://cran.r-project.org/web/packages/glmmTMB/vignettes/troubleshooting.html)

`control=glmmTMBControl(optimizer=optim, optArgs=list(method="BFGS"))`
`control = list(REMLf=FALSE, optimizer = "BFGS"))`

-   [Troubleshooting with glmmTMB](https://cran.r-project.org/web/packages/glmmTMB/vignettes/troubleshooting.html)



## Visualisation and plotting

We can transform a `glmmTMB` model output into a `metafor` object to be able to use the visualisation tools available in `metafor` and `orchard` for example.

[TO BE DONE]

# Simulation study

To assess `glmmTMB` we conducted a simulation study. 

We generated data given four effect size measures:

*   Standardised Mean Difference (SMD)

*   Log Response Ratio (lnRR) 

*   Odd ratio effect sizes (OR)

*   Poisson meta-analysis to model incidence rate ratios (IRR)


## Simulation parameters

| Measure | $\mu$         | $\tau^2$ | $\sigma_T$ | $\sigma_C$ | $\mu_C$ | $n_T$ | $n_C$ | $t_T$ | $t_C$ |
|:--------|:--------------|:--------:|:----------:|:----------:|:-------:|:-----:|:-----:|:-----:|:-----:|
| SMD     | $0.3$         | $0.16$     | $10$       | $10$       | $100$   | $50$  | $50$  |  -    |  -    |
| lnRR    | $\log(1.3)$   | $0.16$     | $10$       | $10$       | $100$   | $50$  | $50$  |  -    |  -    |
| OR      | $\log(1.5)$   | $0.16$     |  -         |  -         |   -     | $50$  | $50$  |  -    |  -    |
| IRR     | $\log(1.5)$   | $0.16$     |  -         |  -         |   -     |  -    |  -    | $50$  | $50$  |




## Results



# References

Bangert-Drowns, R. L., Hurley, M. M., & Wilkinson, B. (2004). _The effects of school-based writing-to-learn interventions on academic achievement: A meta-analysis_. Review of Educational Research, 74, 29-58.

Colditz, G. A., Brewer, T. F., Berkey, C. S., Wilson, M. E., Burdick, E., Fineberg, H. V., & Mosteller, F. (1994). _Efficacy of BCG vaccine in the prevention of tuberculosis: Meta-analysis of the published literature_. Journal of the American Medical Association, 271(9), 698–702. <https://doi.org/10.1001/jama.1994.03510330076038>

Felix, J. A., Stevenson, P. C., & Koricheva, J. (2023). _Plant neighbourhood diversity effects on leaf traits: A meta-analysis_. Functional Ecology, 37(12), 3150–3163. <https://doi.org/10.1111/1365-2435.14441>

Viechtbauer, W. (2010). _Conducting meta-analyses in R with the metafor package_. Journal of Statistical Software, 36(3), 1–48. <https://doi.org/10.18637/jss.v036.i03>

Viechtbauer,W.,&López-López, J.A. (2022). _Location-scale models for meta-analysis_. Research Synthesis Methods, 13(6), 697–715. <https://doi.org/10.1002/jrsm.1562>

White T, Noble D, Senior A, Hamilton W, Viechtbauer W (2022). _metadat: Meta-Analysis Datasets_. R package version 1.2-0, <https://CRAN.R-project.org/package=metadat>.

# Session info

```{r}
#| label: Reproducibility-SessionInfo-R-environment
#| fig-align: "center"
#| out-width: '100%'
#| results: asis
#| message: false
#| warnings: false

library(sessioninfo)
library(details)

si <- session_info()
details(si, summary = 'Current session info', open = FALSE)

```

