---
title: "Fitting meta-analysis with `glmmTMB` package"
author: "Coralie Williams"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    theme: cosmo
    toc-title: "**Table of contents**"
    embed-resources: true
    code-fold: show
    code-tools: true
    number-sections: true
    number-depth: 2
    colorlinks: true
    footnote-location: margin
---

This webpage is supporting material to the paper "*Meta-analysis with `glmmTMB` R package*".

We show how to fit a meta-analysis using the `equalto` covariance structure in `glmmTMB` R package, which allows to specify the sampling error variance-covariance (VCV) matrix.

### Contact {#contact}

For questions or comments, contact Coralie Williams: [coraliewilliams\@outlook.com](mailto:coraliewilliams@outlook.com)



# Fitting a meta-analysis with `glmmTMB`

To demonstrate how to fit a meta-analysis using the `glmmTMB` R package, we will fit a multilevel meta-analytic model as a tutorial example, assuming multiple effect sizes per study. This model synthesises effect sizes across studies by accounting for variation at both the study and observation (effect size) levels, while also modelling correlated residual errors within studies.

Given $j$ effect sizes within $i$ studies, we specify the model as:

$$
y_{ij} = b_0 + u_i + m_{ij} + e_{ij},
$$

where $b_0$ is the overall mean effect, $u_i \sim N(0, \sigma^2_{u}I)$ represents the among-study random effect, $m_{ij} \sim N(0, \sigma^2_{m}I)$ captures the within-study random effect (at the observation-level), and $e_{ij} \sim N(0, \text{VCV})$ are the sampling errors at the observation-level.

$\text{VCV}$ represents a block-diagonal variance-covariance matrix including known (or calculated) sampling variances $\sigma^2_{e}=v_i$ on the diagonal and accounts for correlations among sampling errors from the same study.

### Load `glmmTMB` package

Currently, `equalto` is not on CRAN (yet!), so you need to install `glmmTMB` from the [following branch](https://github.com/coraliewilliams/glmmTMB/tree/equalto_covstruc):

::: callout-note
**Important note:** If you have `glmmTMB` already installed, you will need to uninstall it first.
:::

```{r}
#| eval: false
#| output: false
#| warnings: false

# build package from branch: 
remotes::install_github("coraliewilliams/glmmTMB", ref="equalto_covstruc", subdir="glmmTMB")
```

Load `glmmTMB` package:

```{r}
library(glmmTMB)
```

Check if the `equalto` covariance structure is available (it should be in position 13 of this output) :

```{r}
glmmTMB:::.valid_covstruct
```

### Simulate some data

Now, let's simulate some data following the structure of the above multilevel meta-analysis.

```{r}
#| output: false
#| warnings: false

# load libraries
suppressPackageStartupMessages({
  library(MASS)
  library(kableExtra)
  library(metafor)
})

# set up indices
k.studies <- 10 # number of studies
k.per.study <- 3 # number of effect sizes per study
study <- rep(seq_len(k.studies), times=k.per.study) # study id 
k <- length(study) # total number of effect sizes
id <- seq_len(k) # id of between study effect sizes 
es.id <- unlist(lapply(k.per.study, seq_len)) # id of within study effect sizes


# fixed effect coeff true value (i.e. the overall mean)
b0 <- 0

# simulate sampling errors variances
set.seed(123)
vi <- rbeta(k, 2, 20)

# get VCV matrix assuming within-study correlation of rho=0.5 
V <- matrix(0, nrow = k, ncol = k) 
for (i in 2:k) {
  for (j in 1:i) {
    if (study[i] == study[j]) {
      V[i,j] <- 0.5 * sqrt(vi[i] * vi[j]) 
    }
  }
}
V[upper.tri(V)] <- t(V)[upper.tri(V)] # fill in upper diagonal
diag(V) <- vi # fill in diagonal


# random effects
sigma2.u <- 0.2
sigma2.m <- 0.3
u <- rnorm(k.studies, 0, sqrt(sigma2.u))[study]
m <- rnorm(k, 0, sqrt(sigma2.m))

# simulate multivariate normal sampling errors with specific VCV
set.seed(123)
e <- mvrnorm(n = 1, mu = rep(0, length(vi)), Sigma = V) 
#e <- rnorm(k, 0, sqrt(vi))[study] #without within-study correlation (diag VCV)

# compute y
y <- b0 + u + m + e


# combine into dataframe
dat <- data.frame(y = y,
                  vi = vi,
                  study = study,
                  id = id,
                  es.id = es.id,
                  obs = as.factor(id), #unique ID for propto and equalto (really important it is a factor variable)
                  g = 1 #group ID for propto and equalto (a constant here)
)
```

### Fit `glmmTMB` model

Before we fit the model we need to obtain the within-study error variance-covariance matrix (which accounts for within-study correlation) assuming a constant correlation of $\rho=0.5$. The correlation between effect sizes from different studies is assumed to be zero.

We use the metafor `vcalc` function to obtain a block diagonal VCV matrix for the sampling errors:

```{r}
VCV <- metafor::vcalc(vi, cluster=study, obs=obs, data=dat, rho=0.5)
```

Now we fit the `glmmTMB` model with `equalto` covariance structure.

We specify:

-   the study-level random effect $u_i$ as `(1|study)`.

-   the sampling errors $e_{ij}$ are specified using the covariance structure `equalto(0 + obs|g,VCV)` allowing to specify the VCV matrix of the sampling errors.

-   the within-study random effect $m_{ij}$ that captures the variability at the observation level is modelled as the *residual error* in this model.

::: callout-note
**Important note:** the `equalto` multivariate random effect variable (here it is called `obs`) must be a factor and the corresponding VCV matrix must have rows and columns matching the levels of the `obs` variable.
:::


```{r}
# Fit glmmTMB model with equalto ----------------------------------
fit.equalto <- glmmTMB(y ~ 1 + (1|study) + equalto(0 + obs|g,VCV),
                       data=dat,
                       REML=T)

summary(fit.equalto)
```

### Fit `metafor` model

Let's compare with the `metafor` package `rma.mv` function that fits a multilevel model with random effects. Here the model is specified with two random effects: `(1|study)` corresponding to the among-study ($u_i$) and `(1|id)` corresponding to the within-study ($m_{ij}$) random effects. The sampling errors $e_{ij}$ VCV matrix is specified using the `V` argument.

```{r}
# Fit metafor model --------------------------------------------
fit.metafor <- rma.mv(yi = y, 
                      V = VCV,
                      random = list(~1 | study, ~1 | id),
                      control=list(REMLf=FALSE), #set this to obtain the exact same likelihood as glmmTMB 
                      data = dat) #uses REML by default

summary(fit.metafor)
```

Then, we extract the model results for each model and compare outputs:

```{r}
### Compare output 
metafor <- data.frame(model = "metafor", 
                      est = fit.metafor$b[[1]], 
                      se = fit.metafor$se[[1]], 
                      zval = fit.metafor$zval,
                      pval = fit.metafor$pval,
                      sigma.u = sqrt(fit.metafor$sigma2[1]), ## among study variance
                      sigma.m = sqrt(fit.metafor$sigma2[2])) ## within study variance

equalto <- data.frame(model = "glmmTMB",
                      est = unlist(fixef(fit.equalto))[[1]],
                      se = as.numeric(sqrt(vcov(fit.equalto)[[1]])),
                      zval = summary(fit.equalto)$coefficients$cond[3],
                      pval = summary(fit.equalto)$coefficients$cond[4],
                      sigma.u = exp(fit.equalto$fit$par[[2]]), ##among study variance estimate
                      sigma.m = sigma(fit.equalto)) ##within study variance estimate


output <- rbind(metafor, equalto)
kable(output)

```

The overall mean estimate and random effect variance estimates are identical across the two packages ✅


### Extra analyses

When the number of studies is small (<10) it is preferable to use the $t$-distribution and adjusted degrees of freedom to avoid Type I error rate of coefficient tests and low coverage rates of coefficient confidence intervals ([read more about the recommendations here](https://wviechtb.github.io/metafor/reference/misc-recs.html)).

In `metafor` you can do this directly in the `rma.mv` function when fitting a multilevel/multivariate model with the arguments `test="t"` and `dfs="contain"` (actually if `dfs="contain"` is set this automatically sets `test="t"`).

```{r}
# Fit metafor model --------------------------------------------
fit.metafor.t <- rma.mv(yi = y, 
                      V = VCV,
                      random = list(~1 | study, ~1 | id),
                      test = "t",
                      dfs = "contain", 
                      data = dat) 

summary(fit.metafor.t)
```

With `glmmTMB` you can derive the $t$-test statistic, adjusted degrees of freedom and corresponding P value the following way:

```{r}
# summary stats of fixed effects
fef <- as.data.frame(summary(fit.equalto)$coefficients$cond)
b0.est <- fef$Estimate
b0.se <- fef$`Std. Error`

# if the coefficient of interest varies at the level of corresponding to a random effect (i.e. here it is `study`) then the degree of freedom if the number of unique levels minus the number of estimated coefficients in the model
df.contain <- length(unique(dat$study)) - nrow(fef)
# otherwise it is the total nuber of estimates in the analysis minus the total number of model coefficients
#df.adj <- nrow(dat) - nrow(fef)

# get t-test statistic and p-value
t <- b0.est / b0.se
p <- 2 * pt(abs(t), df = df.contain, lower.tail = FALSE) # two-sided p value
# 95% CI
crit <- qt(0.975, df = df.contain)
ci   <- c(b0.est - crit * b0.se, b0.est + crit * b0.se)

# glmmTMB t-test output
fef.t <- data.frame(Estimate = b0.est,
                    SE = b0.se,
                    tval = t,
                    df = df.contain,
                    pval = p,
                    ci.lb = ci[1],
                    ci.ub = ci[2])
fef.t

```



# Illustrative examples

To illustrate how to fit meta-analyses with `glmmTMB`, we re-analyse three published datasets from distinct research domains:

- **Medicine** — vaccine effectiveness.
- **Ecology & evolution** — plant diversity effects.
- **Social sciences** — school-based learning programmes.

Across these case studies, we show model specification, output, and interpretation using `glmmTMB`.


## Bivariate meta-analysis

Meta-analysis is widely used in medicine to combine evidence across clinical trials and observational studies. As many outcomes are binary (e.g., event vs non-event), common effect sizes are odds ratios, risk ratios, risk differences, and their log transforms.

We will use the Colditz et al. (1994) dataset that is directly avaiable in `metafor` via the `metadat` R package. 
This meta-analysis assessed the effectiveness of the Bacillus Calmette-Guerin (BCG) vaccine against tuberculosis across 13 studies.


```{r}
dat <- dat.colditz1994
```

Reshape to long format so that each arm has its own row (“v” = vaccinated, “c” = unvaccinated/control), and setting control as the reference group.

```{r}
dat_long <- to.long(measure="OR", ai=tpos, bi=tneg, ci=cpos, di=cneg, data=dat.bcg)
levels(dat_long$group) <- c("c", "v")
dat_long$group <- relevel(dat_long$group, ref="c")
```

Then we compute the log-odds effect size measures with `metafor` function `escalc`. We also need to create a constant grouping dummy variable `g` for `glmmTMB` model specification (this needs to be a constant value for all rows). 
```{r}
dat_long <- escalc(measure="PLO", xi=out1, mi=out2, data=dat_long)
dat_long$obs <- as.factor(1:nrow(dat_long))
dat_long$g <- 1 #dummy variable
head(dat_long)
```

Derive the variance-covariance matrix (VCV) with in diagonal the sampling error variances and zero on the off-diagonals (assume studies are independent):

```{r}
VCV <- diag(dat_long$vi)
VCV[1:5,1:5]
```

To fit a bivariate meta-analysis with `glmmTMB` using `equalto` supply the sampling-error variance–covariance matrix (here it is called `VCV`), and fix the dispersion at zero (`dispformula = ~ 0`) since we don't want to model any within-study variance effect. 

```{r}
bv_fit <- glmmTMB(yi ~ group + (0 + group|study) + equalto(0 + obs|g,VCV),
                       dispformula = ~ 0,
                       REML = TRUE,
                       data=dat_long)
```

To obtain the variance component estimates and corresponding correlation $\rho$ (i.e. `rho`) for each group we use the `VarCorr` function:

```{r}
vc <- VarCorr(bv_fit)
var_con <- vc$cond$study["groupc", "groupc"]
var_exp <- vc$cond$study["groupv", "groupv"]
rho.tmb <- vc$cond$study["groupc", "groupv"] / sqrt(var_con * var_exp)
```

Then we can extract the model results as follows: 

```{r}
bv_out <- data.frame(est.b0 = as.numeric(unlist(fixef(bv_fit))[[1]]), 
                     se.b0 = as.numeric(sqrt(vcov(bv_fit)[[1]]))[1], 
                     est.b1 = as.numeric(unlist(fixef(bv_fit))[[2]]), 
                     se.b1 = as.numeric(sqrt(vcov(bv_fit)[[1]]))[4], 
                     tau1 = sqrt(var_con), #study-level heterogeneity for group=control
                     tau2 = sqrt(var_exp), #study-level heterogeneity for group=exp
                     rho = rho.tmb)

bv_out
```



This is equivalent to the following model in `metafor`:

```{r}
bv_fit_rma <- rma.mv(yi, vi, mods = ~ group, 
                     random = ~ group | study, 
                     struct = "UN",
                     data=dat_long)
```

We can extract the model results as follows:

```{r}
bv_out_rma <- with(bv_fit_rma, data.frame(est.b0 = as.numeric(b[1]),
                                      se.b0 = se[1],
                                      est.b1 = as.numeric(b[2]),
                                      se.b1 = se[2],
                                      tau1 = sqrt(tau2[1]),
                                      tau2 = sqrt(tau2[2]),
                                      rho = rho))
bv_out_rma
```




## Phylogenetic meta-analysis

In meta-analysis of ecological and evolutionary studies, effect sizes are often drawn from multiple species. Because species share evolutionary history, these effect size estimates are not independent. To account for this phylogenetic relatedness, we can include a species-level random effect structured by a phylogenetic correlation (which is derived from a phylogenetic tree and assumed model of evolution). We can fit such a random effect in `glmmTMB` with the `propto` covariance structure. 

We will use a published meta-analysis dataset of plant diversity effects on leaf traits (Felix et al., 2023) to showcase how to use `glmmTMB` to fit a phylogenetic meta-analysis.

```{r}
#| eval: false
#| output: false
#| warnings: false


library(readr)
ef <- read_csv("felix_data/effect_sizes.csv")
options(na.action = "na.pass")
head(ef)
```

Data set-up:

```{r}
#| eval: false
#| output: false
#| warnings: false

# load libraries
suppressPackageStartupMessages({
  library(rotl)
  library(ape)
})

## calculate absolute ef
yi_ab <- abs(ef$yi)
ef_ab <- ef
ef_ab$yi <- yi_ab  #converts effect sizes to absolute values

# construct phylogentic tree matrix for use as random factor
species <- unique(ef$species) # list of unique species in meta-analysis
species <- as.character(species) # change to character object
taxa <- tnrs_match_names(species)
tree <- rotl::tol_induced_subtree(taxa$ott_id)
tree$tip.label <- strip_ott_ids(tree$tip.label, remove_underscores = TRUE) # change ids to the names from dataset

# calculate correlations between all species = cor
tree2 <- compute.brlen(tree)
cor <- vcv(tree2, cor = T)

# add phylo random factor
ef$phylo <- ef$species
ef_ab$phylo <- ef_ab$species

### Subgroups for each trait category   

# directional 
phenolics <- subset(ef, trait_category == "phenolics")
LDMC <- subset(ef, trait_category == "LDMC")
C <- subset(ef, trait_category == "C")
N <- subset(ef, trait_category == "N")
terpenes <- subset(ef, trait_category == "terpenes")
thickness <- subset(ef, trait_category == "thickness")
toughness <- subset(ef, trait_category == "toughness")
SLA <- subset(ef, trait_category == "SLA")

# absolute
phenolics_ab <- subset(ef_ab, trait_category == "phenolics")
LDMC_ab <- subset(ef_ab, trait_category == "LDMC")
C_ab <- subset(ef_ab, trait_category == "C")
N_ab <- subset(ef_ab, trait_category == "N")
terpenes_ab <- subset(ef_ab, trait_category == "terpenes")
thickness_ab <- subset(ef_ab, trait_category == "thickness")
toughness_ab <- subset(ef_ab, trait_category == "toughness")
SLA_ab <- subset(ef_ab, trait_category == "SLA")

```

Fit a phylogenetic meta-analysis in `glmmTMB` using two covariance structures:

-   `propto`: to model shared evolutionary history across species
-   `equalto`: to incorporate sampling-errors variance-covariance 


```{r}

```



## Location-scale meta-analysis

We will use a social-science meta-analysis of 48 studies on school-based writing-to-learn interventions. Each study contrasts an intervention group (greater emphasis on writing tasks) with a control group, and study effect size measures are standardised mean differences, where positive values indicate better academic performance under the intervention.


```{r}
dat <- dat.bangertdrowns2004
head(dat)
```







# Simulation study

## Simulation parameters

## Results




# References


Bangert-Drowns, R. L., Hurley, M. M., & Wilkinson, B. (2004). The effects of school-based writing-to-learn interventions on academic achievement: A meta-analysis. Review of Educational Research, 74, 29-58.

Colditz, G. A., Brewer, T. F., Berkey, C. S., Wilson, M. E., Burdick, E., Fineberg, H. V., & Mosteller, F. (1994). Efficacy of BCG vaccine in the prevention of tuberculosis: Meta-analysis of the published literature. Journal of the American Medical Association, 271(9), 698–702. https://doi.org/10.1001/jama.1994.03510330076038


White T, Noble D, Senior A, Hamilton W, Viechtbauer W (2022). _metadat: Meta-Analysis Datasets_. R package version 1.2-0, <https://CRAN.R-project.org/package=metadat>.

Viechtbauer, W. (2010). _Conducting meta-analyses in R with the metafor package_. Journal of Statistical Software, 36(3), 1–48. <https://doi.org/10.18637/jss.v036.i03>


# Session info

```{r}
#| label: Reproducibility-SessionInfo-R-environment
#| fig-align: "center"
#| out-width: '100%'
#| results: asis
#| message: false
#| warnings: false

library(sessioninfo)
library(details)

si <- session_info()
details(si, summary = 'Current session info', open = FALSE)

```

