---
title: "Meta-analysis with glmmTMB R package"
author: "Coralie Williams"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    theme: default
    embed-resources: true
    code-fold: show
    code-tools: true
    number-sections: true
    number-depth: 2
    footnote-location: margin
---

This webpage is supporting material to the paper "*Meta-analysis with `glmmTMB` R package*".

`equalto` covariance structure

### Contact {#contact}

For questions or comments, contact Coralie Williams: [coraliewilliams\@outlook.com](mailto:coraliewilliams@outlook.com)



# Fitting a meta-analysis with `glmmTMB`

To demonstrate how to fit a meta-analysis using the `glmmTMB` R package, we will fit a multilevel meta-analytic model as a tutorial example, assuming multiple effect sizes per study. This model synthesises effect sizes across studies by accounting for variation at both the study and observation (effect size) levels, while also modelling correlated residual errors within studies.

Given $j$ effect sizes within $i$ studies, we specify the model as:

$$
y_{ij} = b_0 + u_i + m_{ij} + e_{ij},
$$

where $b_0$ is the overall mean effect, $u_i \sim N(0, \sigma^2_{u}I)$ represents the among-study random effect, $m_{ij} \sim N(0, \sigma^2_{m}I)$ captures the within-study random effect (at the observation-level), and $e_{ij} \sim N(0, \text{VCV})$ are the sampling errors at the observation-level.

$\text{VCV}$ represents a block-diagonal variance-covariance matrix including known (or calculated) sampling variances $\sigma^2_{e}=v_i$ on the diagonal and accounts for correlations among sampling errors from the same study.

### Load `glmmTMB` package

Currently, `equalto` is not on CRAN (but soon it will be available\*\*\*), so you need to install `glmmTMB` from the [following branch](https://github.com/coraliewilliams/glmmTMB/tree/equalto_covstruc):

::: callout-note
**Important note:** If you have `glmmTMB` already installed, you will need to uninstall it first.
:::

```{r}
#| eval: false
#| output: false
#| warnings: false

# build package from branch: 
remotes::install_github("coraliewilliams/glmmTMB", ref="equalto_covstruc", subdir="glmmTMB")
```

Load `glmmTMB` package:

```{r}
library(glmmTMB)
```

Check if the `equalto` covariance structure is available (it should be in position 13 of this output) :

```{r}
glmmTMB:::.valid_covstruct
```

### Simulate some data

Now, let's simulate some data following the structure of the above multilevel meta-analysis.

```{r}
# load libraries
suppressPackageStartupMessages({
  library(MASS)
  library(kableExtra)
  library(metafor)
})

# set up indices
k.studies <- 50 # number of studies
k.per.study <- 10 # number of effect sizes per study
study <- rep(seq_len(k.studies), times=k.per.study) # study id 
k <- length(study) # total number of effect sizes
id <- seq_len(k) # id of between study effect sizes 
es.id <- unlist(lapply(k.per.study, seq_len)) # id of within study effect sizes


# fixed effect coeff true value (i.e. the overall mean)
b0 <- 0

# simulate sampling errors variances
set.seed(123)
vi <- rbeta(k, 2, 20)

# get VCV matrix assuming within-study correlation of rho=0.5 
V <- matrix(0, nrow = k, ncol = k) 
for (i in 2:k) {
  for (j in 1:i) {
    if (study[i] == study[j]) {
      V[i,j] <- 0.5 * sqrt(vi[i] * vi[j]) 
    }
  }
}
V[upper.tri(V)] <- t(V)[upper.tri(V)] # fill in upper diagonal
diag(V) <- vi # fill in diagonal


# random effects
sigma2.u <- 0.2
sigma2.m <- 0.3
u <- rnorm(k.studies, 0, sqrt(sigma2.u))[study]
m <- rnorm(k, 0, sqrt(sigma2.m))

# simulate multivariate normal sampling errors with specific VCV
set.seed(123)
e <- mvrnorm(n = 1, mu = rep(0, length(vi)), Sigma = V) 
#e <- rnorm(k, 0, sqrt(vi))[study] #without within-study correlation (diag VCV)

# compute y
y <- b0 + u + m + e


# combine into dataframe
dat <- data.frame(y = y,
                  vi = vi,
                  study = study,
                  id = id,
                  es.id = es.id,
                  obs = as.factor(id), #unique ID for propto and equalto (really important it is a factor variable)
                  g = 1 #group ID for propto and equalto (a constant here)
)
```

### Run model

Before we fit the model we need to obtain the within-study error variance-covariance matrix (which accounts for within-study correlation) assuming a constant correlation of $\rho=0.5$. The correlation between effect sizes from different studies is assumed to be zero.

We use the metafor `vcalc` function to obtain a block diagonal VCV matrix for the sampling errors:

```{r}
VCV <- metafor::vcalc(vi, cluster=study, obs=obs, data=dat, rho=0.5)
```

Now we fit the `glmmTMB` model with `equalto` covariance structure.

We specify:

-   the study-level random effect $u_i$ as `(1|study)`.

-   the sampling errors $e_{ij}$ are specified using the covariance structure `equalto(0 + obs|g,VCV)` allowing to specify the VCV matrix of the sampling errors.

-   the within-study random effect $m_{ij}$ that captures the variability at the observation level is modelled as the *residual error* in this model.

```{r}
# Fit glmmTMB model with equalto ----------------------------------
fit.equalto <- glmmTMB(y ~ 1 + (1|study) + equalto(0 + obs|g,VCV),
                       data=dat,
                       REML=T)

```

Let's compare with the `metafor` package `rma.mv` function that fits a multilevel model with random effects. Here the model is specified with two random effects: `(1|study)` corresponding to the among-study ($u_i$) and `(1|id)` corresponding to the within-study ($m_{ij}$) random effects. The sampling errors $e_{ij}$ VCV matrix is specified using the `V` argument.

```{r}
# Fit metafor model --------------------------------------------
fit.metafor <- rma.mv(yi = y, 
                      V = VCV,
                      random = list(~1 | study, ~1 | id),
                      test = "t",
                      dfs = "contain",
                      data = dat) #uses REML by default

fit.metafor
```

Then, we extract the model results for each model and compare outputs:

```{r}
### Compare output 
metafor <- data.frame(model = "metafor", 
                      est = fit.metafor$b[[1]], 
                      se = fit.metafor$se[[1]], 
                      zval = fit.metafor$zval,
                      pval = fit.metafor$pval,
                      sigma.u = sqrt(fit.metafor$sigma2[1]), ## among study variance
                      sigma.m = sqrt(fit.metafor$sigma2[2])) ## within study variance

equalto <- data.frame(model = "equalto",
                      est = unlist(fixef(fit.equalto))[[1]],
                      se = as.numeric(sqrt(vcov(fit.equalto)[[1]])),
                      zval = summary(fit.equalto)$coefficients$cond[3],
                      pval = summary(fit.equalto)$coefficients$cond[4],
                      sigma.u = exp(fit.equalto$fit$par[[2]]), ##among study variance estimate
                      sigma.m = sigma(fit.equalto)) ##within study variance estimate


output <- rbind(metafor, equalto)
kable(output)

```

The overall mean estimate and random effect variance estimates are identical across the two packages.

# Illustrative examples

## Bivariate meta-analysis

```{r}
dat <- dat.colditz1994
head(dat)
```

## Phylogenetic meta-analysis

```{r}
#| eval: false
#| output: false
#| warnings: false


library(readr)
ef <- read_csv("felix_data/effect_sizes.csv")
options(na.action = "na.pass")
head(ef)
```

Data set-up:

```{r}
#| eval: false
#| output: false
#| warnings: false

# load libraries
suppressPackageStartupMessages({
  library(rotl)
  library(ape)
})

## calculate absolute ef
yi_ab <- abs(ef$yi)
ef_ab <- ef
ef_ab$yi <- yi_ab  #converts effect sizes to absolute values

# construct phylogentic tree matrix for use as random factor
species <- unique(ef$species) # list of unique species in meta-analysis
species <- as.character(species) # change to character object
taxa <- tnrs_match_names(species)
tree <- rotl::tol_induced_subtree(taxa$ott_id)
tree$tip.label <- strip_ott_ids(tree$tip.label, remove_underscores = TRUE) # change ids to the names from dataset

# calculate correlations between all species = cor
tree2 <- compute.brlen(tree)
cor <- vcv(tree2, cor = T)

# add phylo random factor
ef$phylo <- ef$species
ef_ab$phylo <- ef_ab$species

### Subgroups for each trait category   

# directional 
phenolics <- subset(ef, trait_category == "phenolics")
LDMC <- subset(ef, trait_category == "LDMC")
C <- subset(ef, trait_category == "C")
N <- subset(ef, trait_category == "N")
terpenes <- subset(ef, trait_category == "terpenes")
thickness <- subset(ef, trait_category == "thickness")
toughness <- subset(ef, trait_category == "toughness")
SLA <- subset(ef, trait_category == "SLA")

# absolute
phenolics_ab <- subset(ef_ab, trait_category == "phenolics")
LDMC_ab <- subset(ef_ab, trait_category == "LDMC")
C_ab <- subset(ef_ab, trait_category == "C")
N_ab <- subset(ef_ab, trait_category == "N")
terpenes_ab <- subset(ef_ab, trait_category == "terpenes")
thickness_ab <- subset(ef_ab, trait_category == "thickness")
toughness_ab <- subset(ef_ab, trait_category == "toughness")
SLA_ab <- subset(ef_ab, trait_category == "SLA")

```

## Location-scale meta-analysis

```{r}
dat <- dat.bangertdrowns2004
head(dat)
```

# Simulation study

## Simulation parameters

## Results
